---
title: "Project: Identification of a Notch transcriptomic signature for breast cancer"
subtitle: "Part 6 - Evaluation of NOTCH signatures in the in vitro validation cohort"
author: "Computational Oncology: Felix Geist"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    highlight: tango
    fig_caption: TRUE
    number_sections: FALSE
    code_folding: hide
    toc: TRUE
    toc_depth: 1
    toc_float:
      collapsed: FALSE
      smooth_scroll: FALSE
bibliography: "`r here::here('src', 'part6.bib')`"
link-citations: true
biblio-style: "apalike"
---

```{r 00_setup, include=FALSE}

knitr::opts_chunk$set(
	fig.height = 4,                  # 4 inch high, 6 inch wide
	fig.width = 6,
	dpi = 300,                       
	dev = c("svg"),                  
	fig.align = "center",
	# fig.path = "../Figures/Part6/",           
	message = FALSE,
	warning = FALSE,
	echo = TRUE,                     
	cache = FALSE                    
)
```

```{css, echo=FALSE}

text {
  font-family: sans-serif;
}

h1, .h1, h2, .h2, h3, .h3 {
    margin-top: 0.5em;
}

div.box { 
  background-color:#EEEEEE; 
  border: 2px solid #0F69AF;
  border-radius: 10px; 
  padding: 1em 1em 1em 4em;
  margin: 1em 0 1em 0;
}

div.obs { 
  background-color:white; 
  border: 2px solid #0F69AF; 
  border-radius: 10px; 
  padding: 1em 1em 1em 4em;
  margin: 1em 0 1em 0;
}

body, td {
   font-size: 14px;
}
code.r{
  font-size: 12px;
}
pre {
  font-size: 14px
}

img {
    max-width: 80%;
}
```

```{r 01-packages, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}

library(tidyverse)

library(readxl)
library(xlsx)
library(openxlsx)
library(knitr)
library(kableExtra)

library(ggplot2)
library(ggpubr)
library(ggrepel)
library(circlize)
library(ggthemes)
library(ComplexHeatmap)
library(scales)
library(cowplot)

library(DESeq2)
library(edgeR)

library(rmarkdown)
library(DT)

library(clusterProfiler)
library(enrichplot)
library(sva)
library(vsn)
library(PCAtools)
library(uwot)
library(pROC)

library(msigdbr)
library(org.Hs.eg.db)

library(coin)
library(rstatix)

library(viridis)

# Merck X-Omics Platform:

library(ProfilerAPI2)
# library(xopdata)
# getsrc <- xopdata::get_src()

# Citations:

knitr::write_bib(file = '../src/part6.bib')

# color schema

merck_colors <- ggthemes::tableau_color_pal("Color Blind")(10)

merck_colors_ext <- colorRamp2(seq(from = 1, to = 40, by = 4), merck_colors[1:10])

subtype_colors <- merck_colors[c(5,9,6)]
names(subtype_colors) <- c("basal", "her2", "luma")

all_subtype_colors <- c(subtype_colors, c("lumb" = as.character(merck_colors[4])))

time_colors <- merck_colors[c(10,7)]
names(time_colors) <- c("8h", "72h")

treatment_colors <- merck_colors[c(1,2,3,4)]
names(treatment_colors) <- c("Notch on", "Notch off", "ground state","Notch on/off")

batch_colors <- merck_colors[c(5,9)]
names(batch_colors) <- c("Validation", "Training")

col_heatmap <- c(merck_colors[1], "white", merck_colors[6])
col_heatmap_not0 <- c(merck_colors[1], merck_colors[3], merck_colors[6])

theme_set(theme_bw())

# heatmap colors:

options(knitr.kable.NA = '')   # This replaces NA in knitr tables to an empty string

```

**Research Questions**

<div class = "box">

We want to identify general applicable signatures for the activation and inactivation of the NOTCH pathway. Therefore, we treated triple negative breast cancer cell lines either with the gamma-secretase inhibitor DAPT or cultivated the cells on immobilized NOTCH ligand Jagged1. Subsequently, the cells were harvested after 8h or 72h of incubation time and their transcriptome analyzed by RNASeq.  

</div>
  
**Methods**

1. Gene Set Enrichment of the published and the optimized Notch signatures using the validation data set
2. Calculation of gene signature scores, therefore, both validation and training data set are combined and batch corrected
3. Correlation of signatures to public data sets to tumor purity / immune infiltration

**Conclusion**  

<div class = "box">

GSEA of the validation cohort in 28 of 56 published signatures as well as in the experimentally derived 20-gene Signature resulted in a significant enrichment. The 20-gene Signature can significantly separate basal-like cell lines from the other subtypes, as well as 38 of the published signatures. 
Only the 20-gene signature significantly, with a large effect size, can separate treatment between Notch On and Notch Off in both the 8h and 72h time point.

</div>

# Methods and Data Sources

## Methods

**Platform and Report**  
The analysis was performed using R `r paste(R.Version()[c("major", "minor")], collapse = ".")` [@R-base] with the extension of the `tidyverse` [@R-tidyverse; @tidyverse2019] (dplyr, forcats, ggplot2, purrr, readr, stringr, tibble, tidyr).   
The report was generated using `Rmarkdown` [@R-rmarkdown; @rmarkdown2018; @rmarkdown2020] and `knitr` [@R-knitr; @knitr2014; @knitr2015].  

**Expression Profiling**  
The RNASeq data had been analzed using `DESeq2` [@R-DESeq2].  

**Visualization**  
Plots are drawn by `ggplot2`, `ggthemes` [@R-ggthemes], `ggpubr` [@R-ggpubr], and `ggrepel` [@R-ggrepel].  
Tables are drawn, using the `knitr` package, together with `kableExtra` styling [@R-kableExtra]. Interactive Tables are integrated using the `DT` package [@R-DT]. Heatmaps are drawn using the `ComplexHeatmap` package [@R-ComplexHeatmap; @ComplexHeatmap2016].     

**Gene set enrichment analysis (GSEA)**  
GSEA was performed by the `fgsea` implementation in `clusterProfiler` [@clusterProfiler2012; @R-clusterProfiler] and visualized by `enrichplot` [@R-enrichplot].  

**Batch Correction**  
To analyze a potential batch effects, the RNASeq data sets are analyzed using `PCAtools` [@R-PCAtools] and `uwot` [@R-uwot], subsequently the batch effect are adjusted for by using an empirical Bayes framework (ComBat), implemented in `sva` [@R-sva].  

**Statistics**   
Statistics were calculated using the `rstatix` package [@R-rstatix].  

**Gene set and genomic annotations**  
In addition to the gene signatures from publications and adaption of gene symbols from mouse to human via `org.Hs.eg.db` and `ClusterProfiler` [@R-AnnotationDbi; @R-org.Hs.eg.db; @clusterProfiler2012; @R-clusterProfiler], all gene signatures with NOTCH in the name were loaded from the MSIGdb via `msigdbr` [@R-msigdbr].  

**ROC Analysis**  
Receiver operating characteristic (ROC) and accuracy rates were calculated with the help of the implementation of `pROC` [@R-pROC; @pROC2011].   
 

```{r 02-load-data}

# Internal Access to Data Lake
# Data sets as described in the manuscript

api = ProfilerAPI2::profiler_api(profile = "myprofile")   ## needed for Rmarkdown connection to ProfilerAPI2
con <- api$conn

### Load all data needed, make the variables more "readable" for comparison:

base::load("../RData/Part1c.RData")
#Training_ddsSE_72h = ddsSE_72h
Training_res_DAPT_72h = res_DAPT_72h
Training_res_Jagged1_72h = res_Jagged1_72h
Training_res_Jagged1_DAPT_72h = res_Jagged1_DAPT_72h
rm(ddsSE_72h, res_DAPT_72h, res_Jagged1_72h, res_Jagged1_DAPT_72h)

base::load("../RData/Part1b.RData")
#Training_ddsSE_8h = ddsSE_8h
Training_res_DAPT_8h = res_DAPT_8h
Training_res_Jagged1_8h = res_Jagged1_8h
Training_res_Jagged1_DAPT_8h = res_Jagged1_DAPT_8h
rm(ddsSE_8h, res_DAPT_8h, res_Jagged1_8h, res_Jagged1_DAPT_8h)

base::load("../RData/Part1a.RData")
Training_TPM = qnt.mat_tpm_agg_log2
#Training_ddsSE_preparation = ddsSE_preparation
#Training_ddsSE = ddsSE
Training_res_DAPT = res_DAPT
Training_res_Jagged1 = res_Jagged1
Training_res_Jagged1_DAPT = res_Jagged1_DAPT
rm(ddsSE, res_DAPT, res_Jagged1, res_Jagged1_DAPT, meta.all_summary, ddsSE_preparation, notch_CCLE_mutations_csv_21Q1_long, qnt.mat_tpm_agg_log2)

base::load("../RData/Part5.RData")

```

# Validation of Signatures

We will validate all published and MSIGDB signatures and all signatures from the optimization approach in the validation cohort experiments by gene set enrichment analysis (GSEA).  

```{r}

### --- Gene set enrichment using fgsea:

validation_all_Jagged1_DAPT <- validation_res_Jagged1_DAPT %>%
  as.data.frame() %>%
  rownames_to_column("symbol") %>%
  dplyr::mutate(stat = -log10(P.Value)*logFC) %>% 
  dplyr::arrange(desc(stat)) %>%
  dplyr::select(-AveExpr, -t, -P.Value, -adj.P.Val, -B, -ensembl, -gene_chrom, -genename, -entrez, -logFC)

validation_all_Jagged1_DAPT_list <- validation_all_Jagged1_DAPT$stat
names(validation_all_Jagged1_DAPT_list) <- validation_all_Jagged1_DAPT$symbol

validation_all_Jagged1_DAPT_8h <- validation_res_Jagged1_DAPT_8h %>%
  as.data.frame() %>%
  rownames_to_column("symbol") %>%
  dplyr::mutate(stat = -log10(P.Value)*logFC) %>% 
  dplyr::arrange(desc(stat)) %>%
  dplyr::select(-AveExpr, -t, -P.Value, -adj.P.Val, -B, -ensembl, -gene_chrom, -genename, -entrez, -logFC)

validation_all_Jagged1_DAPT_8h_list <- validation_all_Jagged1_DAPT_8h$stat
names(validation_all_Jagged1_DAPT_8h_list) <- validation_all_Jagged1_DAPT_8h$symbol

validation_all_Jagged1_DAPT_72h <- validation_res_Jagged1_DAPT_72h %>%
  as.data.frame() %>%
  rownames_to_column("symbol") %>%
  dplyr::mutate(stat = -log10(P.Value)*logFC) %>% 
  dplyr::arrange(desc(stat)) %>%
  dplyr::select(-AveExpr, -t, -P.Value, -adj.P.Val, -B, -ensembl, -gene_chrom, -genename, -entrez, -logFC)

validation_all_Jagged1_DAPT_72h_list <- validation_all_Jagged1_DAPT_72h$stat
names(validation_all_Jagged1_DAPT_72h_list) <- validation_all_Jagged1_DAPT_72h$symbol

rm(validation_res_Jagged1_DAPT_72h,
   validation_res_Jagged1_DAPT_8h,
   validation_res_Jagged1_DAPT)
```

## GSEA of signatures with a good coherence score in TCGA

### GSEA: Validation cohort, comparison Jagged1-DAPT for both time points

```{r, dev='png'}

base::load("../RData/Part2.RData")
rm(cpm_mat_full,
   cs_q,
   cs_df_metabric,
   cs_df_Oslo2,
   cs_df_TCGA,
   metabric_gene_expession_data_scaled_wide_full,
   Oslo2_gene_expession_data_wide_full,
   tpm_mat_full)

### --- Do GSEA

combined_gs.list <- c(gene_set_list_opt_notch)

# select gene sets with CS > 0.2

combined_gs.list <- combined_gs.list[cs_all_optimized$cs_mean_TCGA>0.2 & !is.na(cs_all_optimized$cs_mean_TCGA)]

# interesection of gene sets with genes present in validation DEG

### Jagged1 vs DAPT 8h/72h

gs.list.Validation_all = purrr::map(combined_gs.list, ~intersect(.x, names(validation_all_Jagged1_DAPT_list)))

gs.list.Validation_all <- tibble::tibble(id_geneset_name = names(gs.list.Validation_all),
                                     genes = unname(gs.list.Validation_all),
                                     size = purrr::map_int(gs.list.Validation_all, length),
                                     size.orig = purrr::map_int(combined_gs.list, length))

gs.list.Validation_all_list <- data.frame(gs_name = character(),
                                           gene_symbol = character())

for(i in 1:length(gs.list.Validation_all$id_geneset_name)){
  gs.list.Validation_all_list <- bind_rows(gs.list.Validation_all_list,
                                            data.frame(gs_name = gs.list.Validation_all$id_geneset_name[i],
                                                       gene_symbol = gs.list.Validation_all$genes[[i]]))
  
}

### Jagged1 vs DAPT 8h

gs.list.Validation_8h = purrr::map(combined_gs.list, ~intersect(.x, names(validation_all_Jagged1_DAPT_8h_list)))

gs.list.Validation_8h <- tibble::tibble(id_geneset_name = names(gs.list.Validation_8h),
                                     genes = unname(gs.list.Validation_8h),
                                     size = purrr::map_int(gs.list.Validation_8h, length),
                                     size.orig = purrr::map_int(combined_gs.list, length))

gs.list.Validation_8h_list <- data.frame(gs_name = character(),
                                           gene_symbol = character())

for(i in 1:length(gs.list.Validation_8h$id_geneset_name)){
  gs.list.Validation_8h_list <- bind_rows(gs.list.Validation_8h_list,
                                      data.frame(gs_name = gs.list.Validation_8h$id_geneset_name[i],
                                                 gene_symbol = gs.list.Validation_8h$genes[[i]]))
  
}

### Jagged1 vs 72h

gs.list.Validation_72h = purrr::map(combined_gs.list, ~intersect(.x, names(validation_all_Jagged1_DAPT_72h_list)))

gs.list.Validation_72h <- tibble::tibble(id_geneset_name = names(gs.list.Validation_72h),
                                     genes = unname(gs.list.Validation_72h),
                                     size = purrr::map_int(gs.list.Validation_72h, length),
                                     size.orig = purrr::map_int(combined_gs.list, length))

gs.list.Validation_72h_list <- data.frame(gs_name = character(),
                                           gene_symbol = character())

for(i in 1:length(gs.list.Validation_72h$id_geneset_name)){
  gs.list.Validation_72h_list <- bind_rows(gs.list.Validation_72h_list,
                                      data.frame(gs_name = gs.list.Validation_72h$id_geneset_name[i],
                                                 gene_symbol = gs.list.Validation_72h$genes[[i]]))
  
}

### --- do the GSEA

gsea_Jagged1_DAPT <- GSEA(validation_all_Jagged1_DAPT_list, TERM2GENE = gs.list.Validation_all_list, 
                          pvalueCutoff = 1,
                          minGSSize = 3,
                          maxGSSize = 100)

DT::datatable(gsea_Jagged1_DAPT@result %>%
                dplyr::select(-Description, -core_enrichment) %>% 
                mutate_if(is.numeric, ~signif(., 2)),                       # just use significant numbers
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')        # definition of download buttons
            )
          )

merck_colors_gsea <- colorRamp2(c(-log10(gsea_Jagged1_DAPT@result$p.adjust)[1], 
                                  -log10(0.05),
                                 -log10(gsea_Jagged1_DAPT@result$p.adjust)[nrow(gsea_Jagged1_DAPT@result)]),
                               c(merck_colors[1], "lightgrey", merck_colors[2]))

sign_gsea_Jagged1_DAPT <- gsea_Jagged1_DAPT@result %>% 
  dplyr::filter(p.adjust < 0.05)

```

```{r, fig.width=6, fig.height=4}

if(nrow(as.data.frame(gsea_Jagged1_DAPT)) != 0){
  print(dotplot(gsea_Jagged1_DAPT, 
                x = "NES", 
                size = "setSize",
                showCategory=20,
                font.size = 8) + 
          ggtitle("GSEA of Notch target genesets\n8h+72h Treatment of Breast Cancer Validation Cell lines\nNotch on vs. Notch off") +
          scale_color_viridis(begin = 0, end = 1, option = "D",
                              limits = c(0,0.05), oob=squish))
}

```

### GSEA: Validation cohort, comparison Jagged1-DAPT for the time point: 8h

```{r, dev='png'}

gsea_Jagged1_DAPT_8h <- GSEA(sort(validation_all_Jagged1_DAPT_8h_list, decreasing = TRUE), 
                             TERM2GENE = gs.list.Validation_8h_list, 
                             pvalueCutoff = 1,
                             minGSSize = 3,
                             maxGSSize = 100)

DT::datatable(gsea_Jagged1_DAPT_8h@result %>%
                dplyr::select(-Description, -core_enrichment) %>% 
                mutate_if(is.numeric, ~signif(., 2)),                       # just use significant numbers
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')        # definition of download buttons
            )
          )

merck_colors_gsea <- colorRamp2(c(-log10(gsea_Jagged1_DAPT_8h@result$p.adjust)[1], 
                                  -log10(0.05),
                                 -log10(gsea_Jagged1_DAPT_8h@result$p.adjust)[nrow(gsea_Jagged1_DAPT_8h@result)]),
                               c(merck_colors[1], "lightgrey", merck_colors[2]))
#                                  
# gseaplot2(gsea_Jagged1_DAPT_8h, 
#           geneSetID = gsea_Jagged1_DAPT_8h@result$ID[c(1:3, (nrow(gsea_Jagged1_DAPT_8h@result)-1):nrow(gsea_Jagged1_DAPT_8h@result))], 
#           pvalue_table = FALSE, 
#           color = merck_colors_gsea(-log10(gsea_Jagged1_DAPT_8h@result$p.adjust)[c(1:3, (nrow(gsea_Jagged1_DAPT_8h@result)-1):nrow(gsea_Jagged1_DAPT_8h@result))])[order(gsea_Jagged1_DAPT_8h@result$ID[c(1:3, (nrow(gsea_Jagged1_DAPT_8h@result)-1):nrow(gsea_Jagged1_DAPT_8h@result))])],
#           ES_geom = "line") 

sign_gsea_Jagged1_DAPT_8h <- gsea_Jagged1_DAPT_8h@result %>% 
  dplyr::filter(p.adjust < 0.05)

```


```{r, fig.width=6, fig.height=4}

if(nrow(as.data.frame(gsea_Jagged1_DAPT_8h)) != 0){
  print(dotplot(gsea_Jagged1_DAPT_8h, 
                x = "NES", 
                size = "setSize",
                showCategory=20,
                font.size = 8) + 
          ggtitle("GSEA of Notch target genesets\n8h Treatment of Breast Cancer Cell lines\nNotch on vs. Notch off") +
          scale_color_viridis(begin = 0, end = 1, option = "D",
                              limits = c(0,0.05), oob=squish))
}

```

### GSEA: Validation cohort, comparison Jagged1-DAPT for the time point: 72h

```{r, dev='png'}

gsea_Jagged1_DAPT_72h <- GSEA(sort(validation_all_Jagged1_DAPT_72h_list, decreasing = TRUE), 
                              TERM2GENE = gs.list.Validation_72h_list, 
                              pvalueCutoff = 1,
                          minGSSize = 3,
                          maxGSSize = 100)


DT::datatable(gsea_Jagged1_DAPT@result %>%
                dplyr::select(-Description, -core_enrichment) %>% 
                mutate_if(is.numeric, ~signif(., 2)),                       # just use significant numbers
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')        # definition of download buttons
            )
          )

merck_colors_gsea <- colorRamp2(c(-log10(gsea_Jagged1_DAPT_72h@result$p.adjust)[1], 
                                  -log10(0.05),
                                 -log10(gsea_Jagged1_DAPT_72h@result$p.adjust)[nrow(gsea_Jagged1_DAPT_72h@result)]),
                               c(merck_colors[1], "lightgrey", merck_colors[2]))
#                                  
# gseaplot2(gsea_Jagged1_DAPT_72h, 
#           geneSetID = gsea_Jagged1_DAPT_72h@result$ID[c(1:3, (nrow(gsea_Jagged1_DAPT_72h@result)-1):nrow(gsea_Jagged1_DAPT_72h@result))], 
#           pvalue_table = FALSE, 
#           color = merck_colors_gsea(-log10(gsea_Jagged1_DAPT_72h@result$p.adjust)[c(1:3, (nrow(gsea_Jagged1_DAPT_72h@result)-1):nrow(gsea_Jagged1_DAPT_72h@result))])[order(gsea_Jagged1_DAPT_72h@result$ID[c(1:3, (nrow(gsea_Jagged1_DAPT_72h@result)-1):nrow(gsea_Jagged1_DAPT_72h@result))])],
#           ES_geom = "line") 

sign_gsea_Jagged1_DAPT_72h <- gsea_Jagged1_DAPT_72h@result %>% 
  dplyr::filter(p.adjust < 0.05)

Validation_gsea_summary <- gsea_Jagged1_DAPT@result %>% 
  dplyr::select(ID, NES, p.adjust) %>% 
  dplyr::mutate(gsea = "Jagged1_DAPT") %>% 
  bind_rows(gsea_Jagged1_DAPT_8h@result %>% 
                dplyr::select(ID, NES, p.adjust) %>% 
                dplyr::mutate(gsea = "Jagged1_DAPT_8h")) %>% 
  bind_rows(gsea_Jagged1_DAPT_72h@result %>% 
                dplyr::select(ID, NES, p.adjust) %>% 
                dplyr::mutate(gsea = "Jagged1_DAPT_72h")) %>% 
  as_tibble()

# separate published and optimized signatures:

Validation_gsea.opt_notch_summary <- Validation_gsea_summary %>% 
  dplyr::filter(grepl("Notch 20-Gene Signature", ID))

Validation_gsea.published_notch_summary <- Validation_gsea_summary %>% 
  dplyr::filter(!grepl("Notch 20-Gene Signature", ID))

```


```{r, fig.width=6, fig.height=4}

if(nrow(as.data.frame(gsea_Jagged1_DAPT_72h)) != 0){
  print(dotplot(gsea_Jagged1_DAPT_72h, 
                x = "NES", 
                size = "setSize",
                showCategory=20,
                font.size = 8) + 
          ggtitle("GSEA of Notch target genesets\n72h Treatment of Breast Cancer Cell lines\nNotch on vs. Notch off") +
          scale_color_viridis(begin = 0, end = 1, option = "D",
                              limits = c(0,0.05), oob=squish))
}

```

### Figure: All three comparisons for the three signatures

```{r fig.height=2.5, fig.width=7.2}

Validation_gsea_summary %>% 
  dplyr::mutate(ID = str_replace(ID, "\\_", " "),
                gsea = str_replace(gsea, "Jagged1_DAPT", "Notch On vs. Notch Off"),
                gsea = str_replace(gsea, "_", ": "),
                gsea = str_replace(gsea, "f$", "f: 8h+72h")) %>% 
  dplyr::mutate(gsea = factor(gsea, levels = unique(gsea))) %>% 
  dplyr::mutate(ID = factor(ID, levels = str_sort(unique(.$ID), numeric = TRUE))) %>% 
  ggplot(aes(x = ID,
             y = NES,
             fill = p.adjust)) +
  geom_bar(position="dodge", stat="identity") +
  facet_wrap(~ gsea) +
  scale_fill_gradient(low = merck_colors[6],
                      high = merck_colors[1],
                      limits = c(0,0.05),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90)




Validation_gsea_summary %>% 
  dplyr::mutate(ID = str_replace(ID, "\\_", " "),
                gsea = str_replace(gsea, "Jagged1_DAPT", "Notch On vs. Notch Off"),
                gsea = str_replace(gsea, "_", ": "),
                gsea = str_replace(gsea, "f$", "f: 8h+72h")) %>% 
  dplyr::mutate(gsea = factor(gsea, levels = unique(gsea))) %>% 
  dplyr::mutate(ID = factor(ID, levels = str_sort(unique(.$ID), numeric = TRUE)),
                textlabel = round(NES, digits = 1)) %>% 
  ggplot(aes(x = ID,
             y = NES,
             color = p.adjust,
             label = textlabel)) +
  geom_segment(aes(x = ID,
                   xend = ID,
                   y = 0,
                   yend = NES),
               color = "darkgrey",
               size = 2) +
  geom_point(size = 8,
             shape = 16) +
  geom_text(color = "white",
            fontface = "bold",
            size = 3) +
  facet_wrap(~ gsea) +
  scale_color_gradient(low = merck_colors[6],
                      high = merck_colors[1],
                      limits = c(0,0.05),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90) +
  coord_cartesian(ylim = c(0, 2))




Validation_gsea_summary %>% 
  dplyr::mutate(ID = str_replace(ID, "\\_", " "),
                gsea = str_replace(gsea, "Jagged1_DAPT", "Notch On vs. Notch Off"),
                gsea = str_replace(gsea, "_", ": "),
                gsea = str_replace(gsea, "f$", "f: 8h+72h")) %>% 
  dplyr::mutate(gsea = factor(gsea, levels = unique(gsea))) %>% 
  dplyr::mutate(ID = factor(ID, levels = str_sort(unique(.$ID), numeric = TRUE))) %>% 
  ggplot(aes(x = ID,
             y = NES,
             color = p.adjust)) +
  geom_segment(aes(x = ID,
                   xend = ID,
                   y = 0,
                   yend = NES),
               color = "black") +
  geom_point(size = 8,
             shape = 16) +
  facet_wrap(~ gsea) +
  scale_color_gradient(low = merck_colors[6],
                      high = merck_colors[1],
                      limits = c(0,0.05),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90) +
  coord_cartesian(ylim = c(0, 2))
  

Validation_gsea_summary %>% 
  dplyr::mutate(ID = str_replace(ID, "\\_", " "),
                gsea = str_replace(gsea, "Jagged1_DAPT", "Notch On vs. Notch Off"),
                gsea = str_replace(gsea, "_", ": "),
                gsea = str_replace(gsea, "f$", "f: 8h+72h")) %>% 
  dplyr::mutate(gsea = factor(gsea, levels = unique(gsea))) %>% 
  dplyr::mutate(ID = factor(ID, levels = str_sort(unique(.$ID), numeric = TRUE))) %>% 
  ggplot(aes(x = ID,
             y = NES,
             color = p.adjust)) +
  geom_point(size = 4,
             shape = 16) +
  facet_wrap(~ gsea) +
  scale_color_gradient(low = merck_colors[6],
                      high = merck_colors[1],
                      limits = c(0,0.05),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90) +
  coord_cartesian(ylim = c(0, 2))
  
```

## GSEA of all 15 signatures

### GSEA: Validation cohort, comparison Jagged1-DAPT for both time points

```{r, dev='png'}

#base::load("../RData/04-Notch-Part2-Signatures.RData")
# rm(cpm_mat_full,
#    cs_q,
#    cs_df_metabric,
#    cs_df_Oslo2,
#    cs_df_TCGA,
#    cs_all_optimized,
#    metabric_gene_expession_data_scaled_wide_full,
#    Oslo2_gene_expession_data_wide_full,
#    tpm_mat_full)

### --- Do GSEA

# save all signatures from the initial search
gene_set_list_14_notch <- gene_set_list_opt_notch

combined_gs.list <- c(gene_set_list_opt_notch)

# remove the list with no length

combined_gs.list <- combined_gs.list[!unlist(lapply(combined_gs.list, function(x) all(is.na(x))))]

# interesection of gene sets with genes present in validation DEG

### Jagged1 vs DAPT 8h/72h

gs.list.Validation_all = purrr::map(combined_gs.list, ~intersect(.x, names(validation_all_Jagged1_DAPT_list)))

gs.list.Validation_all <- tibble::tibble(id_geneset_name = names(gs.list.Validation_all),
                                     genes = unname(gs.list.Validation_all),
                                     size = purrr::map_int(gs.list.Validation_all, length),
                                     size.orig = purrr::map_int(combined_gs.list, length))

gs.list.Validation_all_list <- data.frame(gs_name = character(),
                                           gene_symbol = character())

for(i in 1:length(gs.list.Validation_all$id_geneset_name)){
  gs.list.Validation_all_list <- bind_rows(gs.list.Validation_all_list,
                                            data.frame(gs_name = gs.list.Validation_all$id_geneset_name[i],
                                                       gene_symbol = gs.list.Validation_all$genes[[i]]))
  
}

### Jagged1 vs DAPT 8h

gs.list.Validation_8h = purrr::map(combined_gs.list, ~intersect(.x, names(validation_all_Jagged1_DAPT_8h_list)))

gs.list.Validation_8h <- tibble::tibble(id_geneset_name = names(gs.list.Validation_8h),
                                     genes = unname(gs.list.Validation_8h),
                                     size = purrr::map_int(gs.list.Validation_8h, length),
                                     size.orig = purrr::map_int(combined_gs.list, length))

gs.list.Validation_8h_list <- data.frame(gs_name = character(),
                                           gene_symbol = character())

for(i in 1:length(gs.list.Validation_8h$id_geneset_name)){
  gs.list.Validation_8h_list <- bind_rows(gs.list.Validation_8h_list,
                                      data.frame(gs_name = gs.list.Validation_8h$id_geneset_name[i],
                                                 gene_symbol = gs.list.Validation_8h$genes[[i]]))
  
}

### Jagged1 vs 72h

gs.list.Validation_72h = purrr::map(combined_gs.list, ~intersect(.x, names(validation_all_Jagged1_DAPT_72h_list)))

gs.list.Validation_72h <- tibble::tibble(id_geneset_name = names(gs.list.Validation_72h),
                                     genes = unname(gs.list.Validation_72h),
                                     size = purrr::map_int(gs.list.Validation_72h, length),
                                     size.orig = purrr::map_int(combined_gs.list, length))

gs.list.Validation_72h_list <- data.frame(gs_name = character(),
                                           gene_symbol = character())

for(i in 1:length(gs.list.Validation_72h$id_geneset_name)){
  gs.list.Validation_72h_list <- bind_rows(gs.list.Validation_72h_list,
                                      data.frame(gs_name = gs.list.Validation_72h$id_geneset_name[i],
                                                 gene_symbol = gs.list.Validation_72h$genes[[i]]))
  
}

### --- do the GSEA

gsea_Jagged1_DAPT <- GSEA(validation_all_Jagged1_DAPT_list, TERM2GENE = gs.list.Validation_all_list, 
                          pvalueCutoff = 1,
                          minGSSize = 3,
                          maxGSSize = 100)

DT::datatable(gsea_Jagged1_DAPT@result %>%
                dplyr::select(-Description, -core_enrichment) %>% 
                mutate_if(is.numeric, ~signif(., 2)),                       # just use significant numbers
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')        # definition of download buttons
            )
          )

merck_colors_gsea <- colorRamp2(c(-log10(gsea_Jagged1_DAPT@result$p.adjust)[1], 
                                  -log10(0.05),
                                 -log10(gsea_Jagged1_DAPT@result$p.adjust)[nrow(gsea_Jagged1_DAPT@result)]),
                               c(merck_colors[1], "lightgrey", merck_colors[2]))

sign_gsea_Jagged1_DAPT <- gsea_Jagged1_DAPT@result %>% 
  dplyr::filter(p.adjust < 0.05)

```

```{r, fig.width=5, fig.height=5}

if(nrow(as.data.frame(gsea_Jagged1_DAPT)) != 0){
  print(dotplot(gsea_Jagged1_DAPT, 
                x = "NES", 
                size = "setSize",
                showCategory=20,
                font.size = 8) + 
          ggtitle("GSEA of Notch target genesets\n8h+72h Treatment of Breast Cancer Validation Cell lines\nNotch on vs. Notch off") +
          scale_color_viridis(begin = 0, end = 1, option = "D",
                              limits = c(0,0.05), oob=squish))
}

```

### GSEA: Validation cohort, comparison Jagged1-DAPT for the time point: 8h

```{r, dev='png'}

gsea_Jagged1_DAPT_8h <- GSEA(sort(validation_all_Jagged1_DAPT_8h_list, decreasing = TRUE), 
                             TERM2GENE = gs.list.Validation_8h_list, 
                             pvalueCutoff = 1,
                             minGSSize = 3,
                             maxGSSize = 100)

DT::datatable(gsea_Jagged1_DAPT_8h@result %>%
                dplyr::select(-Description, -core_enrichment) %>% 
                mutate_if(is.numeric, ~signif(., 2)),                       # just use significant numbers
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')        # definition of download buttons
            )
          )

merck_colors_gsea <- colorRamp2(c(-log10(gsea_Jagged1_DAPT_8h@result$p.adjust)[1], 
                                  -log10(0.05),
                                 -log10(gsea_Jagged1_DAPT_8h@result$p.adjust)[nrow(gsea_Jagged1_DAPT_8h@result)]),
                               c(merck_colors[1], "lightgrey", merck_colors[2]))
#                                  
# gseaplot2(gsea_Jagged1_DAPT_8h, 
#           geneSetID = gsea_Jagged1_DAPT_8h@result$ID[c(1:3, (nrow(gsea_Jagged1_DAPT_8h@result)-1):nrow(gsea_Jagged1_DAPT_8h@result))], 
#           pvalue_table = FALSE, 
#           color = merck_colors_gsea(-log10(gsea_Jagged1_DAPT_8h@result$p.adjust)[c(1:3, (nrow(gsea_Jagged1_DAPT_8h@result)-1):nrow(gsea_Jagged1_DAPT_8h@result))])[order(gsea_Jagged1_DAPT_8h@result$ID[c(1:3, (nrow(gsea_Jagged1_DAPT_8h@result)-1):nrow(gsea_Jagged1_DAPT_8h@result))])],
#           ES_geom = "line") 

sign_gsea_Jagged1_DAPT_8h <- gsea_Jagged1_DAPT_8h@result %>% 
  dplyr::filter(p.adjust < 0.05)

```


```{r, fig.width=5, fig.height=5}

if(nrow(as.data.frame(gsea_Jagged1_DAPT_8h)) != 0){
  print(dotplot(gsea_Jagged1_DAPT_8h, 
                x = "NES", 
                size = "setSize",
                showCategory=20,
                font.size = 8) + 
          ggtitle("GSEA of Notch target genesets\n8h Treatment of Breast Cancer Cell lines\nNotch on vs. Notch off") +
          scale_color_viridis(begin = 0, end = 1, option = "D",
                              limits = c(0,0.05), oob=squish))
}

```

### GSEA: Validation cohort, comparison Jagged1-DAPT for the time point: 72h

```{r, dev='png'}

gsea_Jagged1_DAPT_72h <- GSEA(sort(validation_all_Jagged1_DAPT_72h_list, decreasing = TRUE), 
                              TERM2GENE = gs.list.Validation_72h_list, 
                              pvalueCutoff = 1,
                          minGSSize = 3,
                          maxGSSize = 100)


DT::datatable(gsea_Jagged1_DAPT@result %>%
                dplyr::select(-Description, -core_enrichment) %>% 
                mutate_if(is.numeric, ~signif(., 2)),                       # just use significant numbers
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')        # definition of download buttons
            )
          )

merck_colors_gsea <- colorRamp2(c(-log10(gsea_Jagged1_DAPT_72h@result$p.adjust)[1], 
                                  -log10(0.05),
                                 -log10(gsea_Jagged1_DAPT_72h@result$p.adjust)[nrow(gsea_Jagged1_DAPT_72h@result)]),
                               c(merck_colors[1], "lightgrey", merck_colors[2]))
#                                  
# gseaplot2(gsea_Jagged1_DAPT_72h, 
#           geneSetID = gsea_Jagged1_DAPT_72h@result$ID[c(1:3, (nrow(gsea_Jagged1_DAPT_72h@result)-1):nrow(gsea_Jagged1_DAPT_72h@result))], 
#           pvalue_table = FALSE, 
#           color = merck_colors_gsea(-log10(gsea_Jagged1_DAPT_72h@result$p.adjust)[c(1:3, (nrow(gsea_Jagged1_DAPT_72h@result)-1):nrow(gsea_Jagged1_DAPT_72h@result))])[order(gsea_Jagged1_DAPT_72h@result$ID[c(1:3, (nrow(gsea_Jagged1_DAPT_72h@result)-1):nrow(gsea_Jagged1_DAPT_72h@result))])],
#           ES_geom = "line") 

sign_gsea_Jagged1_DAPT_72h <- gsea_Jagged1_DAPT_72h@result %>% 
  dplyr::filter(p.adjust < 0.05)

Validation_gsea_summary <- gsea_Jagged1_DAPT@result %>% 
  dplyr::select(ID, NES, p.adjust) %>% 
  dplyr::mutate(gsea = "Jagged1_DAPT") %>% 
  bind_rows(gsea_Jagged1_DAPT_8h@result %>% 
                dplyr::select(ID, NES, p.adjust) %>% 
                dplyr::mutate(gsea = "Jagged1_DAPT_8h")) %>% 
  bind_rows(gsea_Jagged1_DAPT_72h@result %>% 
                dplyr::select(ID, NES, p.adjust) %>% 
                dplyr::mutate(gsea = "Jagged1_DAPT_72h")) %>% 
  as_tibble()

# separate published and optimized signatures:

Validation_gsea.opt_notch_summary <- Validation_gsea_summary %>% 
  dplyr::filter(grepl("Notch 20-Gene Signature", ID))

Validation_gsea.published_notch_summary <- Validation_gsea_summary %>% 
  dplyr::filter(!grepl("Notch 20-Gene Signature", ID))

```


```{r, fig.width=5, fig.height=5}

if(nrow(as.data.frame(gsea_Jagged1_DAPT_72h)) != 0){
  print(dotplot(gsea_Jagged1_DAPT_72h, 
                x = "NES", 
                size = "setSize",
                showCategory=20,
                font.size = 8) + 
          ggtitle("GSEA of Notch target genesets\n72h Treatment of Breast Cancer Cell lines\nNotch on vs. Notch off") +
          scale_color_viridis(begin = 0, end = 1, option = "D",
                              limits = c(0,0.05), oob=squish))
}

```


### Figure: All three comparisons for the 15 signatures

```{r fig.height=2.5, fig.width=10}

Validation_gsea_summary %>% 
  dplyr::mutate(ID = str_replace(ID, "\\_", " "),
                gsea = str_replace(gsea, "Jagged1_DAPT", "Notch On vs. Notch Off"),
                gsea = str_replace(gsea, "_", ": "),
                gsea = str_replace(gsea, "f$", "f: 8h+72h")) %>% 
  dplyr::mutate(gsea = factor(gsea, levels = unique(gsea))) %>% 
  dplyr::mutate(ID = factor(ID, levels = str_sort(unique(.$ID), numeric = TRUE))) %>% 
  ggplot(aes(x = ID,
             y = NES,
             fill = p.adjust)) +
  geom_bar(position="dodge", stat="identity") +
  facet_wrap(~ gsea) +
  scale_fill_gradient(low = merck_colors[6],
                      high = merck_colors[1],
                      limits = c(0,0.05),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90)


Validation_gsea_summary %>% 
  dplyr::mutate(ID = str_replace(ID, "\\_", " "),
                gsea = str_replace(gsea, "Jagged1_DAPT", "Notch On vs. Notch Off"),
                gsea = str_replace(gsea, "_", ": "),
                gsea = str_replace(gsea, "f$", "f: 8h+72h")) %>% 
  dplyr::mutate(gsea = factor(gsea, levels = unique(gsea))) %>% 
  dplyr::mutate(ID = factor(ID, levels = str_sort(unique(.$ID), numeric = TRUE)),
                textlabel = round(NES, digits = 1)) %>% 
  ggplot(aes(x = ID,
             y = NES,
             color = p.adjust,
             label = textlabel)) +
  geom_segment(aes(x = ID,
                   xend = ID,
                   y = 0,
                   yend = NES),
               color = "darkgrey",
               size = 1) +
  geom_point(size = 6,
             shape = 16) +
  geom_text(color = "white",
            fontface = "bold",
            size = 2) +
  facet_wrap(~ gsea) +
  scale_color_gradient(low = merck_colors[6],
                      high = merck_colors[1],
                      limits = c(0,0.05),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90) +
  coord_cartesian(ylim = c(-2, 2))


Validation_gsea_summary %>% 
  dplyr::mutate(ID = str_replace(ID, "\\_", " "),
                gsea = str_replace(gsea, "Jagged1_DAPT", "Notch On vs. Notch Off"),
                gsea = str_replace(gsea, "_", ": "),
                gsea = str_replace(gsea, "f$", "f: 8h+72h")) %>% 
  dplyr::mutate(gsea = factor(gsea, levels = unique(gsea))) %>% 
  dplyr::mutate(ID = factor(ID, levels = str_sort(unique(.$ID), numeric = TRUE))) %>% 
  ggplot(aes(x = ID,
             y = NES,
             color = p.adjust)) +
  geom_segment(aes(x = ID,
                   xend = ID,
                   y = 0,
                   yend = NES),
               color = "black") +
  geom_point(size = 3,
             shape = 16) +
  facet_wrap(~ gsea) +
  scale_color_gradient(low = merck_colors[6],
                      high = merck_colors[1],
                      limits = c(0,0.05),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90) +
  coord_cartesian(ylim = c(-2, 2))
  

Validation_gsea_summary %>% 
  dplyr::mutate(ID = str_replace(ID, "\\_", " "),
                gsea = str_replace(gsea, "Jagged1_DAPT", "Notch On vs. Notch Off"),
                gsea = str_replace(gsea, "_", ": "),
                gsea = str_replace(gsea, "f$", "f: 8h+72h")) %>% 
  dplyr::mutate(gsea = factor(gsea, levels = unique(gsea))) %>% 
  dplyr::mutate(ID = factor(ID, levels = str_sort(unique(.$ID), numeric = TRUE))) %>% 
  ggplot(aes(x = ID,
             y = NES,
             color = p.adjust)) +
  geom_point(size = 3,
             shape = 16) +
  facet_wrap(~ gsea) +
  scale_color_gradient(low = merck_colors[6],
                      high = merck_colors[1],
                      limits = c(0,0.05),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90) +
  coord_cartesian(ylim = c(-2, 2))




```

## GSEA: Final Signature + MSIGDB Signatures

```{r}

base::load("../RData/Part3.RData")

# signatures to show:

focus_signatures <- c("Notch 20-Gene Signature", sort(c( 
"Braune_2016_TNBC_Notch_sig_CSL_KO",
"Castel_2013_C2C12_Dll1_vs_DAPT_targets_24h_up",
"WP_NOTCH_SIGNALING_PATHWAY",
"GOBP_NOTCH_SIGNALING_INVOLVED_IN_HEART_DEVELOPMENT",
"Labban_2018_SCC_RBPJ_targets",
"Leontovich_2018_NOTCH3_nuclear_reprogramming_BRCA",
"PID_NOTCH_PATHWAY",
"Stoeck2014_TNBC_ACC_GSI_response_1",
"WP_CANONICAL_AND_NONCANONICAL_NOTCH_SIGNALING",
"KEGG_NOTCH_SIGNALING_PATHWAY",
"Poulsen_2018_N1_repressed",
"HALLMARK_NOTCH_SIGNALING",
"BIOCARTA_NOTCH_PATHWAY",
"VILIMAS_NOTCH1_TARGETS_UP",
"GOBP_POSITIVE_REGULATION_OF_NOTCH_SIGNALING_PATHWAY",
"Xu_2017_MDAMB231_RBPJ_targets")))
```


### GSEA: Validation cohort, comparison Jagged1-DAPT for both time points

```{r, dev='png'}

### --- Do GSEA

combined_gs.list <- c(gs.list.opt_notch, gs.list.published_notch)

# interesection of gene sets with genes present in validation DEG

### Jagged1 vs DAPT 8h/72h

gs.list.Validation_all = purrr::map(combined_gs.list, ~intersect(.x, names(validation_all_Jagged1_DAPT_list)))

gs.list.Validation_all <- tibble::tibble(id_geneset_name = names(gs.list.Validation_all),
                                     genes = unname(gs.list.Validation_all),
                                     size = purrr::map_int(gs.list.Validation_all, length),
                                     size.orig = purrr::map_int(combined_gs.list, length))

gs.list.Validation_all_list <- data.frame(gs_name = character(),
                                           gene_symbol = character())

for(i in 1:length(gs.list.Validation_all$id_geneset_name)){
  gs.list.Validation_all_list <- bind_rows(gs.list.Validation_all_list,
                                            data.frame(gs_name = gs.list.Validation_all$id_geneset_name[i],
                                                       gene_symbol = gs.list.Validation_all$genes[[i]]))
  
}

### Jagged1 vs DAPT 8h

gs.list.Validation_8h = purrr::map(combined_gs.list, ~intersect(.x, names(validation_all_Jagged1_DAPT_8h_list)))

gs.list.Validation_8h <- tibble::tibble(id_geneset_name = names(gs.list.Validation_8h),
                                     genes = unname(gs.list.Validation_8h),
                                     size = purrr::map_int(gs.list.Validation_8h, length),
                                     size.orig = purrr::map_int(combined_gs.list, length))

gs.list.Validation_8h_list <- data.frame(gs_name = character(),
                                           gene_symbol = character())

for(i in 1:length(gs.list.Validation_8h$id_geneset_name)){
  gs.list.Validation_8h_list <- bind_rows(gs.list.Validation_8h_list,
                                      data.frame(gs_name = gs.list.Validation_8h$id_geneset_name[i],
                                                 gene_symbol = gs.list.Validation_8h$genes[[i]]))
  
}

### Jagged1 vs 72h

gs.list.Validation_72h = purrr::map(combined_gs.list, ~intersect(.x, names(validation_all_Jagged1_DAPT_72h_list)))

gs.list.Validation_72h <- tibble::tibble(id_geneset_name = names(gs.list.Validation_72h),
                                     genes = unname(gs.list.Validation_72h),
                                     size = purrr::map_int(gs.list.Validation_72h, length),
                                     size.orig = purrr::map_int(combined_gs.list, length))

gs.list.Validation_72h_list <- data.frame(gs_name = character(),
                                           gene_symbol = character())

for(i in 1:length(gs.list.Validation_72h$id_geneset_name)){
  gs.list.Validation_72h_list <- bind_rows(gs.list.Validation_72h_list,
                                      data.frame(gs_name = gs.list.Validation_72h$id_geneset_name[i],
                                                 gene_symbol = gs.list.Validation_72h$genes[[i]]))
  
}

### --- do the GSEA

gsea_Jagged1_DAPT <- GSEA(validation_all_Jagged1_DAPT_list, TERM2GENE = gs.list.Validation_all_list, 
                          pvalueCutoff = 1,
                          minGSSize = 3,
                          maxGSSize = 100)

DT::datatable(gsea_Jagged1_DAPT@result %>%
                dplyr::select(-Description, -core_enrichment) %>% 
                mutate_if(is.numeric, ~signif(., 2)),                       # just use significant numbers
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')        # definition of download buttons
            )
          )

merck_colors_gsea <- colorRamp2(c(-log10(gsea_Jagged1_DAPT@result$p.adjust)[1], 
                                  -log10(0.05),
                                 -log10(gsea_Jagged1_DAPT@result$p.adjust)[nrow(gsea_Jagged1_DAPT@result)]),
                               c(merck_colors[1], "lightgrey", merck_colors[2]))
                                 
# gseaplot2(gsea_Jagged1_DAPT, 
#           geneSetID = gsea_Jagged1_DAPT@result$ID[c(1:3, (nrow(gsea_Jagged1_DAPT@result)-1):nrow(gsea_Jagged1_DAPT@result))], 
#           pvalue_table = FALSE, 
#           color = merck_colors_gsea(-log10(gsea_Jagged1_DAPT@result$p.adjust)[c(1:3, (nrow(gsea_Jagged1_DAPT@result)-1):nrow(gsea_Jagged1_DAPT@result))])[order(gsea_Jagged1_DAPT@result$ID[c(1:3, (nrow(gsea_Jagged1_DAPT@result)-1):nrow(gsea_Jagged1_DAPT@result))])],
#           ES_geom = "line") 

sign_gsea_Jagged1_DAPT <- gsea_Jagged1_DAPT@result %>% 
  dplyr::filter(p.adjust < 0.05)

```

```{r, fig.width=5, fig.height=5}

if(nrow(as.data.frame(gsea_Jagged1_DAPT)) != 0){
  print(dotplot(gsea_Jagged1_DAPT %>% 
                  dplyr::filter(ID %in% !!focus_signatures), 
                x = "NES", 
                size = "setSize",
                showCategory=20,
                font.size = 8) + 
          ggtitle("GSEA of Notch target genesets\n8h+72h Treatment of Breast Cancer Validation Cell lines\nNotch on vs. Notch off") +
          scale_color_viridis(begin = 0, end = 1, option = "D",
                              limits = c(0,0.05), oob=squish))
}

```

### GSEA: Validation cohort, comparison Jagged1-DAPT for the time point: 8h

```{r, dev='png'}

gsea_Jagged1_DAPT_8h <- GSEA(sort(validation_all_Jagged1_DAPT_8h_list, decreasing = TRUE), 
                             TERM2GENE = gs.list.Validation_8h_list, 
                             pvalueCutoff = 1,
                             minGSSize = 3,
                             maxGSSize = 100)

DT::datatable(gsea_Jagged1_DAPT_8h@result %>%
                dplyr::select(-Description, -core_enrichment) %>% 
                mutate_if(is.numeric, ~signif(., 2)),                       # just use significant numbers
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')        # definition of download buttons
            )
          )

merck_colors_gsea <- colorRamp2(c(-log10(gsea_Jagged1_DAPT_8h@result$p.adjust)[1], 
                                  -log10(0.05),
                                 -log10(gsea_Jagged1_DAPT_8h@result$p.adjust)[nrow(gsea_Jagged1_DAPT_8h@result)]),
                               c(merck_colors[1], "lightgrey", merck_colors[2]))
#                                  
# gseaplot2(gsea_Jagged1_DAPT_8h, 
#           geneSetID = gsea_Jagged1_DAPT_8h@result$ID[c(1:3, (nrow(gsea_Jagged1_DAPT_8h@result)-1):nrow(gsea_Jagged1_DAPT_8h@result))], 
#           pvalue_table = FALSE, 
#           color = merck_colors_gsea(-log10(gsea_Jagged1_DAPT_8h@result$p.adjust)[c(1:3, (nrow(gsea_Jagged1_DAPT_8h@result)-1):nrow(gsea_Jagged1_DAPT_8h@result))])[order(gsea_Jagged1_DAPT_8h@result$ID[c(1:3, (nrow(gsea_Jagged1_DAPT_8h@result)-1):nrow(gsea_Jagged1_DAPT_8h@result))])],
#           ES_geom = "line") 

sign_gsea_Jagged1_DAPT_8h <- gsea_Jagged1_DAPT_8h@result %>% 
  dplyr::filter(p.adjust < 0.05)

```


```{r, fig.width=5, fig.height=5}

if(nrow(as.data.frame(gsea_Jagged1_DAPT_8h)) != 0){
  print(dotplot(gsea_Jagged1_DAPT_8h %>% 
                  dplyr::filter(ID %in% !!focus_signatures), 
                x = "NES", 
                size = "setSize",
                showCategory=20,
                font.size = 8) + 
          ggtitle("GSEA of Notch target genesets\n8h Treatment of Breast Cancer Cell lines\nNotch on vs. Notch off") +
          scale_color_viridis(begin = 0, end = 1, option = "D",
                              limits = c(0,0.05), oob=squish))
}

```

### GSEA: Validation cohort, comparison Jagged1-DAPT for the time point: 72h

```{r, dev='png'}

gsea_Jagged1_DAPT_72h <- GSEA(sort(validation_all_Jagged1_DAPT_72h_list, decreasing = TRUE), 
                              TERM2GENE = gs.list.Validation_72h_list, 
                              pvalueCutoff = 1,
                          minGSSize = 3,
                          maxGSSize = 100)


DT::datatable(gsea_Jagged1_DAPT@result %>%
                dplyr::select(-Description, -core_enrichment) %>% 
                mutate_if(is.numeric, ~signif(., 2)),                       # just use significant numbers
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')        # definition of download buttons
            )
          )

merck_colors_gsea <- colorRamp2(c(-log10(gsea_Jagged1_DAPT_72h@result$p.adjust)[1], 
                                  -log10(0.05),
                                 -log10(gsea_Jagged1_DAPT_72h@result$p.adjust)[nrow(gsea_Jagged1_DAPT_72h@result)]),
                               c(merck_colors[1], "lightgrey", merck_colors[2]))
#                                  
# gseaplot2(gsea_Jagged1_DAPT_72h, 
#           geneSetID = gsea_Jagged1_DAPT_72h@result$ID[c(1:3, (nrow(gsea_Jagged1_DAPT_72h@result)-1):nrow(gsea_Jagged1_DAPT_72h@result))], 
#           pvalue_table = FALSE, 
#           color = merck_colors_gsea(-log10(gsea_Jagged1_DAPT_72h@result$p.adjust)[c(1:3, (nrow(gsea_Jagged1_DAPT_72h@result)-1):nrow(gsea_Jagged1_DAPT_72h@result))])[order(gsea_Jagged1_DAPT_72h@result$ID[c(1:3, (nrow(gsea_Jagged1_DAPT_72h@result)-1):nrow(gsea_Jagged1_DAPT_72h@result))])],
#           ES_geom = "line") 

sign_gsea_Jagged1_DAPT_72h <- gsea_Jagged1_DAPT_72h@result %>% 
  dplyr::filter(p.adjust < 0.05)

Validation_gsea_summary <- gsea_Jagged1_DAPT@result %>% 
  dplyr::select(ID, NES, p.adjust) %>% 
  dplyr::mutate(gsea = "Jagged1_DAPT") %>% 
  bind_rows(gsea_Jagged1_DAPT_8h@result %>% 
                dplyr::select(ID, NES, p.adjust) %>% 
                dplyr::mutate(gsea = "Jagged1_DAPT_8h")) %>% 
  bind_rows(gsea_Jagged1_DAPT_72h@result %>% 
                dplyr::select(ID, NES, p.adjust) %>% 
                dplyr::mutate(gsea = "Jagged1_DAPT_72h")) %>% 
  as_tibble()

# separate published and optimized signatures:

Validation_gsea.opt_notch_summary <- Validation_gsea_summary %>% 
  dplyr::filter(grepl("Notch 20-Gene Signature", ID))

Validation_gsea.published_notch_summary <- Validation_gsea_summary %>% 
  dplyr::filter(!grepl("Notch 20-Gene Signature", ID))

```


```{r, fig.width=5, fig.height=5}

if(nrow(as.data.frame(gsea_Jagged1_DAPT_72h)) != 0){
  print(dotplot(gsea_Jagged1_DAPT_72h %>% 
                  dplyr::filter(ID %in% !!focus_signatures), 
                x = "NES", 
                size = "setSize",
                showCategory=20,
                font.size = 8) + 
          ggtitle("GSEA of Notch target genesets\n72h Treatment of Breast Cancer Cell lines\nNotch on vs. Notch off") +
          scale_color_viridis(begin = 0, end = 1, option = "D",
                              limits = c(0,0.05), oob=squish))
}

```


### Figure: All three comparisons for the tested signatures

```{r fig.height=12, fig.width=10}
Validation_gsea_summary %>% 
  dplyr::mutate(ID = factor(ID, levels = Validation_gsea_summary %>%
                                      dplyr::filter(gsea == "Jagged1_DAPT") %>% 
                                      dplyr::arrange(desc(NES)) %>% 
                                      pull(ID),
                            labels = Validation_gsea_summary %>%
                              dplyr::filter(gsea == "Jagged1_DAPT") %>% 
                              dplyr::arrange(desc(NES)) %>% 
                              pull(ID) %>% 
                              str_replace_all("\\_", " ")),
                gsea = str_replace(gsea, "Jagged1_DAPT", "Notch On vs. Notch Off"),
                gsea = str_replace(gsea, "_", ": "),
                gsea = str_replace(gsea, "f$", "f: 8h+72h")) %>% 
  dplyr::mutate(gsea = factor(gsea, levels = unique(gsea))) %>% 
  ggplot(aes(x = ID,
             y = NES,
             fill = p.adjust)) +
  geom_bar(position="dodge", stat="identity") +
  facet_wrap( ~ gsea, ncol = 1) +
  scale_fill_gradient(low = merck_colors[6],
                      high = merck_colors[1],
                      limits = c(0,0.05),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90)
```


```{r fig.height=12, fig.width=12}
Validation_gsea_summary %>% 
  dplyr::mutate(ID = factor(ID, levels = Validation_gsea_summary %>%
                                      dplyr::filter(gsea == "Jagged1_DAPT") %>% 
                                      dplyr::arrange(desc(NES)) %>% 
                                      pull(ID),
                            labels = Validation_gsea_summary %>%
                              dplyr::filter(gsea == "Jagged1_DAPT") %>% 
                              dplyr::arrange(desc(NES)) %>% 
                              pull(ID) %>% 
                              str_replace_all("\\_", " ")),
                gsea = str_replace(gsea, "Jagged1_DAPT", "Notch On vs. Notch Off"),
                gsea = str_replace(gsea, "_", ": "),
                gsea = str_replace(gsea, "f$", "f: 8h+72h")) %>% 
  dplyr::mutate(gsea = factor(gsea, levels = unique(gsea))) %>% 
  dplyr::mutate(textlabel = round(NES, digits = 1)) %>% 
  ggplot(aes(x = ID,
             y = NES,
             color = p.adjust,
             label = textlabel)) +
  geom_segment(aes(x = ID,
                   xend = ID,
                   y = 0,
                   yend = NES),
               color = "darkgrey",
               size = 1) +
  geom_point(size = 5,
             shape = 16) +
  geom_text(color = "white",
            fontface = "bold",
            size = 2) +
  facet_wrap(~ gsea, ncol = 1) +
  scale_color_gradient(low = merck_colors[6],
                      high = merck_colors[1],
                      limits = c(0,0.05),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90) +
  coord_cartesian(ylim = c(-2, 2))

```


```{r fig.height=10, fig.width=5}
Validation_gsea_summary %>% 
  dplyr::filter(ID %in% !!focus_signatures) %>% 
  dplyr::mutate(ID = factor(ID, levels = Validation_gsea_summary %>% 
                              dplyr::filter(ID %in% !!focus_signatures) %>%
                              dplyr::filter(gsea == "Jagged1_DAPT") %>% 
                              dplyr::arrange(desc(NES)) %>% 
                              dplyr::pull(ID),
                            labels = Validation_gsea_summary %>% 
                              dplyr::filter(ID %in% !!focus_signatures) %>%
                              dplyr::filter(gsea == "Jagged1_DAPT") %>% 
                              dplyr::arrange(desc(NES)) %>% 
                              dplyr::pull(ID) %>% 
                              str_replace_all("\\_", " ")),
                gsea = str_replace(gsea, "Jagged1_DAPT", "Notch On vs. Notch Off"),
                gsea = str_replace(gsea, "_", ": "),
                gsea = str_replace(gsea, "f$", "f: 8h+72h")) %>% 
  dplyr::mutate(gsea = factor(gsea, levels = unique(gsea))) %>% 
  ggplot(aes(x = ID,
             y = NES,
             fill = p.adjust)) +
  geom_bar(position="dodge", stat="identity") +
  facet_wrap( ~ gsea, ncol = 1) +
  scale_fill_gradient(low = merck_colors[6],
                      high = merck_colors[1],
                      limits = c(0,0.05),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90)
```


```{r fig.height=10, fig.width=6}
Validation_gsea_summary %>% 
  dplyr::filter(ID %in% !!focus_signatures) %>% 
  dplyr::mutate(ID = factor(ID, levels = Validation_gsea_summary %>% 
                              dplyr::filter(ID %in% !!focus_signatures) %>%
                              dplyr::filter(gsea == "Jagged1_DAPT") %>% 
                              dplyr::arrange(desc(NES)) %>% 
                              dplyr::pull(ID),
                            labels = Validation_gsea_summary %>% 
                              dplyr::filter(ID %in% !!focus_signatures) %>%
                              dplyr::filter(gsea == "Jagged1_DAPT") %>% 
                              dplyr::arrange(desc(NES)) %>% 
                              dplyr::pull(ID) %>% 
                              str_replace_all("\\_", " ")),
                gsea = str_replace(gsea, "Jagged1_DAPT", "Notch On vs. Notch Off"),
                gsea = str_replace(gsea, "_", ": "),
                gsea = str_replace(gsea, "f$", "f: 8h+72h")) %>% 
  dplyr::mutate(gsea = factor(gsea, levels = unique(gsea))) %>% 
  dplyr::mutate(textlabel = round(NES, digits = 1)) %>% 
  ggplot(aes(x = ID,
             y = NES,
             color = p.adjust,
             label = textlabel)) +
  geom_segment(aes(x = ID,
                   xend = ID,
                   y = 0,
                   yend = NES),
               color = "darkgrey",
               size = 1) +
  geom_point(size = 5,
             shape = 16) +
  geom_text(color = "white",
            fontface = "bold",
            size = 2) +
  facet_wrap(~ gsea, ncol = 1) +
  scale_color_gradient(low = merck_colors[6],
                      high = merck_colors[1],
                      limits = c(0,0.05),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90) +
  coord_cartesian(ylim = c(-2, 2))


```


# Cell Line Characteristica

## Expression data of breast cancer cell lines

Signature calculation will be done on the scaled TPM values. To understand whether there are batch effects between the TPM values of the two experimental conditions, both the training and the validation cohort are combined and batch corrected using Combat.  

```{r, dev='png'}
Validation_df <- validation_TPM %>% 
  dplyr::left_join(validation_subtypes,
                   by = "condition_cell_line") %>%
  dplyr::select(sample_name, stripped_cell_line_name, gene_symbol, quant_tpm,
                condition_treatment, condition_time, condition_subtype) %>% 
  dplyr::mutate(condition_replicate = "1",
                condition_batch = "Validation")

Training_df <- Training_TPM %>% 
  dplyr::select(sample_name, stripped_cell_line_name, gene_symbol = symbol, quant_tpm,
                condition_treatment, condition_time, condition_subtype, condition_replicate) %>% 
  dplyr::mutate(condition_batch = "Training")

tpm_df <- bind_rows(Validation_df,
                    Training_df)

tpm_meta <- tpm_df %>% 
  dplyr::select(-quant_tpm, -gene_symbol) %>% 
  dplyr::distinct()

tpm_df_wide <- tpm_df %>% 
  dplyr::select(sample_name, quant_tpm, gene_symbol) %>% 
  dplyr::filter(!is.na(gene_symbol)) %>% 
  dplyr::distinct() %>% 
  dplyr::group_by(sample_name, gene_symbol) %>% 
  dplyr::summarise(quant_tpm = sum(quant_tpm)) %>% 
  dplyr::ungroup() %>% 
  tidyr::pivot_wider(names_from = sample_name,
                     values_from = quant_tpm) %>% 
  column_to_rownames("gene_symbol")

tpm_meta <- tpm_meta[match(colnames(tpm_df_wide), tpm_meta$sample_name),]

# exclude genes that have NAs

tpm_df_wide_na <- tpm_df_wide %>% 
  na.omit()

tpm_mat_scaled <- tpm_df_wide_na %>%
  t() %>% 
  as.data.frame() %>% 
  dplyr::mutate(across(where(is.numeric), ~as.vector(scale(log2(.x + 1))))) %>% 
  t() %>% 
  na.omit()

# PCA

library(FactoMineR)

PCA(tpm_mat_scaled, scale.unit = TRUE, ncp = 5, graph = TRUE)

p <- pca(tpm_mat_scaled, 
         metadata = (tpm_meta %>% 
           column_to_rownames("sample_name")), 
         removeVar = 0.1)

biplot(p,
       title = "PC1 vs PC2",
       subtitle = "Breast Cancer experimental batches",
       lab = NA,
       pointSize = 2,
       colby = 'condition_batch',
       #shape = "Treatment",
       colkey = as.character(merck_colors),
       hline = 0, vline = 0,
       legendPosition = 'right') +
  ggrepel::geom_text_repel(aes(label = p$metadata$stripped_cell_line_name))

```

## Batch Correction

```{r, dev='png'}
### --- Batch Correction
## remove genes with no counts and run ComBat on what's left

batch_k <- rowSums(tpm_df_wide_na) > 5

batch_tpm_k <- tpm_df_wide_na %>% 
  as.data.frame() %>% 
  dplyr::filter(batch_k)

tpm_meta <- tpm_meta[match(colnames(batch_tpm_k), tpm_meta$sample_name),]

batch_tpm <- sva::ComBat(as.matrix(batch_tpm_k),
                         as.factor(tpm_meta$condition_batch))

batch_tpm[batch_tpm < 0] <- 0

## put the zero genes back in
m <- matrix(rep(0, ncol(batch_tpm) * sum(!batch_k)),
            ncol = ncol(batch_tpm), nrow = sum(!batch_k))

rownames(m) <- rownames(tpm_df_wide_na)[!batch_k]
batch_tpm <- rbind(batch_tpm,
                   m)

batch_tpm_scaled <- batch_tpm %>%
  t() %>% 
  as.data.frame() %>% 
  dplyr::mutate(across(where(is.numeric), ~as.vector(scale(log2(.x + 1))))) %>% 
  t() %>% 
  na.omit()

p <- pca(batch_tpm_scaled, 
         metadata = (tpm_meta %>% 
           column_to_rownames("sample_name")), 
         removeVar = 0.1)

biplot(p,
       title = "PC1 vs PC2",
       subtitle = "Batch corrected Breast Cancer experimental batches",
       lab = NA,
       pointSize = 2,
       colby = 'condition_batch',
       #shape = "Treatment",
       colkey = as.character(merck_colors),
       hline = 0, vline = 0,
       legendPosition = 'right') +
  ggrepel::geom_text_repel(aes(label = p$metadata$stripped_cell_line_name))

biplot(p,
       title = "PC1 vs PC2",
       subtitle = "Batch corrected Breast Cancer experimental batches",
       lab = NA,
       pointSize = 2,
       colby = 'condition_subtype',
       #shape = "Treatment",
       colkey = as.character(merck_colors),
       hline = 0, vline = 0,
       legendPosition = 'right') +
  ggrepel::geom_text_repel(aes(label = p$metadata$stripped_cell_line_name))

biplot(p,
       title = "PC1 vs PC2",
       subtitle = "Batch corrected Breast Cancer cell lines",
       lab = NA,
       pointSize = 2,
       colby = 'stripped_cell_line_name',
       #shape = "Treatment",
       colkey = merck_colors_ext(1:40),
       hline = 0, vline = 0,
       legendPosition = 'right') +
  ggrepel::geom_text_repel(aes(label = p$metadata$stripped_cell_line_name))
### --- UMAP
```

<div class = "box">

Batch correction was not needed, as no batch has been found in the TPM data set.  

</div>

## UMAP

```{r, dev='png', fig.height=4, fig.width=5}
z_umap <- umap(t(batch_tpm_scaled), pca = 10)

tpm_meta_outlier_umap <- tpm_meta %>% 
  column_to_rownames("sample_name")

z_umap_df <- data.frame(umap1 = z_umap[,1],
           umap2 = z_umap[,2],
           tpm_meta_outlier_umap,
           stringsAsFactors = FALSE)

rownames(z_umap_df) <- rownames(tpm_meta_outlier_umap)

ggscatter(z_umap_df,
          x = "umap1",
          y = "umap2",
          color = "condition_subtype",
          shape = "condition_time",
          size = 3,
          legend = "right",
          palette = merck_colors,
          title = "Dimensionality Reduction with UMAP") +
  guides(col = guide_legend(nrow = 14)) +
  geom_text_repel(data = z_umap_df %>% group_by(stripped_cell_line_name, condition_subtype) %>% 
               summarise(umap1 = mean(umap1), 
                         umap2 = mean(umap2)),
             aes(x = umap1,
                 y = umap2,
                 label = stripped_cell_line_name,
                 color = condition_subtype))


ggscatter(z_umap_df,
          x = "umap1",
          y = "umap2",
          color = "condition_batch",
          shape = "condition_time",
          size = 3,
          legend = "right",
          palette = merck_colors,
          title = "Dimensionality Reduction with UMAP") +
  guides(col = guide_legend(nrow = 14)) +
  geom_text_repel(data = z_umap_df %>% group_by(stripped_cell_line_name, condition_batch) %>% 
               summarise(umap1 = mean(umap1), 
                         umap2 = mean(umap2)),
             aes(x = umap1,
                 y = umap2,
                 label = stripped_cell_line_name,
                 color = condition_batch))



# prepare long data frame:

batch_tpm_long <- batch_tpm %>% 
  as.data.frame() %>% 
  rownames_to_column("id_gene_symbol") %>% 
  tidyr::pivot_longer(-1,
                      names_to = "sample_name",
                      values_to = "quant_tpm_tmm") %>% 
  dplyr::mutate(quant_log2tpm = log2(quant_tpm_tmm+1)) %>% 
  dplyr::group_by(id_gene_symbol) %>% 
  dplyr::mutate(quant_scaled_tpm = as.numeric(scale(quant_log2tpm))) %>% 
  dplyr::ungroup()

batch_tpm_long <- batch_tpm_long %>% 
  dplyr::left_join(tpm_meta)

batch_tpm_long <- batch_tpm_long %>% 
  dplyr::mutate(condition_treatment = ifelse(condition_treatment == "FC + DAPT", "DAPT", condition_treatment))


# scaled tpm data:

tpm_df_scaled <- tpm_df %>%
  dplyr::rename(id_gene_symbol = gene_symbol) %>%
  dplyr::group_by(id_gene_symbol) %>% 
  dplyr::mutate(quant_scaled_tpm = as.numeric(scale(quant_tpm))) %>% 
  dplyr::ungroup()

## save for app:

save(tpm_df_scaled, file = "../RData/BRCA_Cell_Lines_TPM.RData")
save(batch_tpm_long, file = "../RData/BRCA_Cell_Lines_TPM_batch_corrected.RData")

rm(batch_tpm_long, batch_tpm, batch_tpm_scaled,tpm_df_wide_na, Validation_df, Training_df, Training_TPM, tpm_df, batch_tpm_k)

```

## Cell line characteristica

```{r 08e-receptors-expression-hm, fig.height=2.5, fig.width=5}

# load wikipathway notch

Wikipathways_Notch <- msigdbr(species = "Homo sapiens") %>%
  dplyr::filter(gs_subcat == "CP:WIKIPATHWAYS",
                gs_name == "WP_NOTCH_SIGNALING")

notch_pathways_genes_symbol <- unique(Wikipathways_Notch$human_gene_symbol) 
notch_pathways_genes_entrez <- unique(Wikipathways_Notch$human_entrez_gene) 

### CCLE phenotype data

id_depmap_sanger_phe_cdr_dem <- "depmap_sanger_phe_cdr_dem_20q1_prod_budzinska_2"

all_ccle_cell_lines <- api$conn %>% 
  tbl(id_depmap_sanger_phe_cdr_dem) %>% 
  pull(id_cell_line_name) %>% 
  unique()

# table(unique(batch_tpm_long$stripped_cell_line_name) %in% all_ccle_cell_lines)
# FALSE  TRUE 
#     1    22 

# unique(batch_tpm_long$stripped_cell_line_name)[!unique(batch_tpm_long$stripped_cell_line_name) %in% all_ccle_cell_lines]
# KPL4

# cell line data:

depmap_phe <- api$conn %>% 
  tbl(id_depmap_sanger_phe_cdr_dem) %>% 
   dplyr::filter(id_cell_line_name %in% !!unique(tpm_df_scaled$stripped_cell_line_name)) %>% 
   dplyr::select(id_cell_line_name, id_depmap_id) %>% 
   collect()

# depmap_phe$id_cell_line_name %>% unique

tpm_df_scaled <- tpm_df_scaled %>% 
  dplyr::left_join(depmap_phe,
                   by = c("stripped_cell_line_name" = "id_cell_line_name"))

### mutation data

# get first, which cell lines are covered by wes:

id_depmap_ccle_dna_wes_snv <- "depmap_ccle_dna_wes_snv_20q4_dev_kreis_1"

cell_lines_depmap <- api$conn %>% 
  tbl(id_depmap_ccle_dna_wes_snv) %>% 
  dplyr::filter(id_depmap_id %in% !!unique(tpm_df_scaled$id_depmap_id)) %>% 
  dplyr::pull(id_depmap_id) %>% 
  unique()

notch_mutation_data <- api$conn %>% 
  tbl(id_depmap_ccle_dna_wes_snv) %>% 
  dplyr::filter(id_depmap_id %in% !!unique(tpm_df_scaled$id_depmap_id),
                id_gene_entrez %in% !!unique(notch_pathways_genes_entrez),
                variant_annotation != "silent") %>% 
  dplyr::collect()

# if variant_protein is NA -> take variant_genome and variant_consequence

notch_mutation_data$variant <- ifelse(is.na(notch_mutation_data$variant_protein), 
                                               paste0(notch_mutation_data$variant_consequence,"(", notch_mutation_data$variant_genome, ")"),
                                               notch_mutation_data$variant_protein)

notch_CCLE_mutations_20q4 <- notch_mutation_data %>% 
  dplyr::select(c("variant", "id_gene_symbol", "id_depmap_id"))

# Aggregate multiple mutations per gene into one string:

notch_CCLE_mutations_20q4_agg <- aggregate(variant ~ id_gene_symbol + id_depmap_id, 
                                              data = notch_CCLE_mutations_20q4, 
                                              FUN = function(x) paste(unique(x), collapse=", "))


notch_CCLE_mutations_20q4_wide <- pivot_wider(notch_CCLE_mutations_20q4_agg,
                                      names_from = id_gene_symbol,
                                      names_prefix = "mutated_",
                                      values_from = variant)

# fill with cell lines:
notch_CCLE_mutations_20q4_wide <- data.frame(id_depmap_id  = cell_lines_depmap) %>% 
  dplyr::left_join(notch_CCLE_mutations_20q4_wide)

# replace NA

notch_CCLE_mutations_20q4_wide <- notch_CCLE_mutations_20q4_wide %>%
  mutate_at(vars(matches("mutated_")), function(x) ifelse(is.na(x), "wt", x))

# add now to the "real used ones"

notch_CCLE_mutations_20q4_wide <- data.frame(id_depmap_id = unique(tpm_df_scaled$id_depmap_id),
           stringsAsFactors = FALSE) %>%
  left_join(notch_CCLE_mutations_20q4_wide,
            by = "id_depmap_id")

notch_CCLE_mutations_20q4_long <- notch_CCLE_mutations_20q4_wide %>%
  pivot_longer(-id_depmap_id,
               names_to = "id_gene_symbol",
               values_to = "mutation",
               names_prefix = "mutated_") %>%
  left_join(tpm_df_scaled %>%
              dplyr::select(stripped_cell_line_name, id_depmap_id) %>%
              dplyr::distinct())



```

```{r 08f-receptors-expression-hm, fig.height=8, fig.width=12}

# DO NOT USE BATCH CORRECTED DATA:

receptor_expression <- tpm_df_scaled %>%
  filter(condition_treatment == "FC") %>%
  filter(id_gene_symbol %in% c("ERBB2","ERBB3","ESR1","PGR","AR")) %>%
  dplyr::select(id_gene_symbol, quant_scaled_tpm, condition_cell_line = stripped_cell_line_name, condition_subtype, condition_time, condition_replicate, condition_batch) %>%
  arrange(id_gene_symbol, condition_subtype, condition_cell_line, desc(condition_time))

matrix_receptor_expression <- receptor_expression %>% 
  dplyr::select(-condition_subtype, -condition_replicate) %>%
  tidyr::pivot_wider(names_from = c(condition_cell_line, condition_time, condition_batch),
                     values_from = quant_scaled_tpm,
                     values_fn = mean) %>%
  column_to_rownames("id_gene_symbol")

# notch_expression <- batch_tpm_long %>%
notch_expression <- tpm_df_scaled %>%
  filter(condition_treatment == "FC") %>%
  filter(id_gene_symbol %in% c("NOTCH1", "NOTCH2", "NOTCH3", "NOTCH4")) %>%
  dplyr::select(id_gene_symbol, quant_scaled_tpm, condition_cell_line = stripped_cell_line_name, condition_subtype, condition_time, condition_replicate, condition_batch) %>%
  arrange(id_gene_symbol, condition_subtype, condition_cell_line, desc(condition_time))

matrix_notch_expression <- notch_expression %>% 
  dplyr::select(-condition_subtype, -condition_replicate) %>%
  tidyr::pivot_wider(names_from = c(condition_cell_line, condition_batch, condition_time),
                     values_from = quant_scaled_tpm,
                     values_fn = mean) %>%
  column_to_rownames("id_gene_symbol")


notch_CCLE_mutations_use_wide <- notch_expression %>%
  dplyr::select(-id_gene_symbol, -quant_scaled_tpm) %>%
  dplyr::distinct() %>%
  left_join(notch_CCLE_mutations_20q4_long %>%
    dplyr::select(-id_depmap_id),
    by = c("condition_cell_line" = "stripped_cell_line_name")) %>%
  arrange(id_gene_symbol, condition_subtype, condition_cell_line, desc(condition_time))
  

xlsx::write.xlsx(notch_CCLE_mutations_use_wide %>%
                   dplyr::select(-condition_time, -condition_replicate, -condition_batch) %>%
                   dplyr::distinct() %>%
                   pivot_wider(names_from = id_gene_symbol,
                               values_from = mutation),
                 file = "../results/NOTCH_pathway_mutations.xlsx")

matrix_notch_mutation <- notch_expression %>% 
  dplyr::select(-condition_subtype, -id_gene_symbol, -quant_scaled_tpm) %>%
  dplyr::left_join(notch_CCLE_mutations_use_wide) %>% 
  dplyr::select(-condition_subtype, -condition_replicate) %>%
  dplyr::distinct() %>%
  tidyr::pivot_wider(names_from = c(condition_cell_line, condition_batch, condition_time),
                     values_from = mutation) %>%
  dplyr::filter(!is.na(id_gene_symbol)) %>%
  column_to_rownames("id_gene_symbol")

# header

Subtype <- as.character(receptor_expression[,c("condition_cell_line", "condition_subtype", "condition_batch", "condition_time")] %>% distinct() %>% pull("condition_subtype"))
Subtype_col <- all_subtype_colors[unique(Subtype)]

Timepoint <- as.character(receptor_expression[,c("condition_cell_line", "condition_time", "condition_batch", "condition_subtype")] %>% distinct() %>% pull("condition_time"))
Timepoint_col <- time_colors[unique(Timepoint)]

Batch <- as.character(receptor_expression[,c("condition_cell_line", "condition_time", "condition_batch", "condition_subtype")] %>% distinct() %>% pull("condition_batch"))
Batch_col <- batch_colors[unique(Batch)]

Cell_Line <- as.character(receptor_expression[,c("condition_cell_line", "condition_time", "condition_batch", "condition_subtype")] %>% distinct() %>% pull("condition_cell_line"))
set.seed(1234)

Cell_Line_col <- sample(merck_colors_ext(1:40))[1:length(unique(Cell_Line))]
names(Cell_Line_col) <- unique(Cell_Line)

col_fun = colorRamp2(c(-2, 0, 2), col_heatmap)

col_ha = ComplexHeatmap::columnAnnotation(Subtype = Subtype,
                                          `Cell Line` = Cell_Line,
                                          Timepoint = Timepoint,
                                          Batch = Batch,
                                          col = list(Subtype = Subtype_col,
                                                     `Cell Line` = Cell_Line_col,
                                                     Timepoint = Timepoint_col,
                                                     Batch = Batch_col), 
                                          #annotation_name_side = "left",
                                          annotation_legend_param = list(
                                            Subtype = list(direction = "horizontal")
                                            )
)

ht <- ComplexHeatmap::Heatmap(matrix_receptor_expression,
                        cluster_rows = FALSE,
                        cluster_columns = FALSE,
                       # column_split  = col_split,
                        row_title = "Expression of\nEndocrine\nReceptors",
                        row_title_rot = 0,
                        row_title_gp = gpar(fontsize = 10),
                        top_annotation = col_ha,
                        show_column_names = TRUE,
                        col = col_fun,
                        rect_gp = gpar(col = "white", lwd = 0.5),
                        heatmap_legend_param = list(
                          title = "scaled log2(TPM+1)", 
                          #title_position = "leftcenter-rot",
                          direction = "horizontal")
)

ht2 <- ComplexHeatmap::Heatmap(matrix_notch_expression,
                        cluster_rows = FALSE,
                        cluster_columns = FALSE,
                       # column_split  = col_split,
                        row_title = "Expression of\nNotch\nReceptors",
                        row_title_rot = 0,
                        row_title_gp = gpar(fontsize = 10),
                        show_column_names = TRUE,
                        col = col_fun,
                        rect_gp = gpar(col = "white", lwd = 0.5),
                        show_heatmap_legend = FALSE,
                        heatmap_legend_param = list(
                          title = "scaled log2(TPM+1)", 
                          title_position = "leftcenter-rot")
)

mutation_names <- unique(unlist(matrix_notch_mutation))
mutation_names <- mutation_names[!is.na(mutation_names)]

col_mut = c("lightgrey",sample(merck_colors_ext(1:40))[1:(length(mutation_names)-1)])
names(col_mut) <- mutation_names


ht3 <- ComplexHeatmap::Heatmap(matrix_notch_mutation,
                        cluster_rows = FALSE,
                        cluster_columns = FALSE,
                       # column_split  = col_split,
                        row_title = "Mutations in\nNotch\nReceptors",
                        row_title_rot = 0,
                        row_title_gp = gpar(fontsize = 10),
                        column_names_gp = gpar(fontsize = 6),
                        show_column_names = TRUE,
                        col = col_mut,
                        na_col = "white",
                        rect_gp = gpar(col = "white", lwd = 0.5),
                        heatmap_legend_param = list(
                          title = "NOTCH Pathway\nMutations")
                        
)

draw(ht %v% ht2 %v% ht3, heatmap_legend_side = "right", annotation_legend_side = "right", merge_legends = TRUE)

```

```{r 09-notch-expression-hm, fig.height=7, fig.width=10}

receptor_expression <- tpm_df_scaled %>%
  filter(condition_treatment == "FC") %>%
  filter(id_gene_symbol %in% c("ERBB2","ERBB3","ESR1","PGR","AR")) %>%
  dplyr::select(id_gene_symbol, quant_scaled_tpm, condition_cell_line = stripped_cell_line_name, condition_subtype, condition_time, condition_replicate, condition_batch) %>%
  arrange(id_gene_symbol, condition_subtype, condition_cell_line, desc(condition_time))

matrix_receptor_expression <- receptor_expression %>% 
  dplyr::select(-condition_subtype, -condition_replicate, -condition_batch, -condition_time) %>%
  tidyr::pivot_wider(names_from = c(condition_cell_line),
                     values_from = quant_scaled_tpm,
                     values_fn = mean) %>%
  column_to_rownames("id_gene_symbol")


notch_expression <- tpm_df_scaled %>%
  filter(condition_treatment == "FC") %>%
  filter(id_gene_symbol %in% c("NOTCH1", "NOTCH2", "NOTCH3", "NOTCH4")) %>%
  dplyr::select(id_gene_symbol, quant_scaled_tpm, condition_cell_line = stripped_cell_line_name, condition_subtype, condition_time, condition_replicate, condition_batch) %>%
  arrange(id_gene_symbol, condition_subtype, condition_cell_line, desc(condition_time))

matrix_notch_expression <- notch_expression %>% 
  dplyr::select(-condition_subtype, -condition_replicate, -condition_batch, -condition_time) %>%
  tidyr::pivot_wider(names_from = c(condition_cell_line),
                     values_from = quant_scaled_tpm,
                     values_fn = mean) %>%
  column_to_rownames("id_gene_symbol")

notch_CCLE_mutations_use_wide <- notch_expression %>%
  dplyr::select(-id_gene_symbol, -quant_scaled_tpm, -condition_time, -condition_replicate, -condition_batch) %>%
  dplyr::distinct() %>%
  left_join(notch_CCLE_mutations_20q4_long %>%
    dplyr::select(-id_depmap_id),
    by = c("condition_cell_line" = "stripped_cell_line_name")) %>%
  arrange(id_gene_symbol, condition_subtype, condition_cell_line)
  
matrix_notch_mutation <- notch_expression %>% 
  dplyr::select(-condition_subtype, -id_gene_symbol, -quant_scaled_tpm,
                -condition_time, -condition_replicate, -condition_batch) %>%
  dplyr::left_join(notch_CCLE_mutations_use_wide) %>% 
  dplyr::select(-condition_subtype) %>%
  dplyr::distinct() %>%
  tidyr::pivot_wider(names_from = c(condition_cell_line),
                     values_from = mutation) %>%
  dplyr::filter(!is.na(id_gene_symbol)) %>%
  column_to_rownames("id_gene_symbol")

matrix_notch_mutation <- matrix_notch_mutation %>% 
  dplyr::mutate(across(everything(), ~ ifelse(. == "wt", "wt", paste0(!!rownames(matrix_notch_mutation), ": ", .))))

# header

Subtype <- as.character(receptor_expression[,c("condition_cell_line", "condition_subtype")] %>% distinct() %>% pull("condition_subtype"))
Subtype_col <- all_subtype_colors[unique(Subtype)]

Cell_Line <- as.character(receptor_expression[,c("condition_cell_line", "condition_subtype")] %>% distinct() %>% pull("condition_cell_line"))
set.seed(1234)

Cell_Line_col <- sample(merck_colors_ext(1:40))[1:length(unique(Cell_Line))]
names(Cell_Line_col) <- unique(Cell_Line)


Batch <- receptor_expression[,c("condition_cell_line", "condition_batch")] %>%
                        distinct() %>%
    tidyr::pivot_wider(names_from = "condition_cell_line",
                       values_from = "condition_batch",
                       values_fn = function(x) paste(sort(x), collapse = " + "))
                        
Batch_ordered <- as.character(Batch)
names(Batch_ordered) <- names(Batch)

Batch_ordered <- Batch_ordered[match(Cell_Line, names(Batch_ordered))]
Batch_ordered <- as.character(Batch_ordered)

Batch_col <- merck_colors[5:(4+length(unique(Batch)))]
names(Batch_col) <- unique(Batch)


col_fun = colorRamp2(c(-2, 0, 2), col_heatmap)

col_ha = ComplexHeatmap::columnAnnotation(Subtype = Subtype,
                                          `Cell Line` = Cell_Line,
                                          Batch = Batch_ordered,
                                          col = list(Subtype = Subtype_col,
                                                     `Cell Line` = Cell_Line_col,
                                                     Batch = Batch_col), 
                                          annotation_legend_param = list(
                                            Subtype = list(direction = "horizontal")
                                            )
)

ht <- ComplexHeatmap::Heatmap(matrix_receptor_expression,
                        cluster_rows = FALSE,
                        cluster_columns = FALSE,
                       # column_split  = col_split,
                        row_title = "Expression of\nEndocrine\nReceptors",
                        row_title_rot = 0,
                        row_title_gp = gpar(fontsize = 10),
                        top_annotation = col_ha,
                        show_column_names = TRUE,
                        col = col_fun,
                        rect_gp = gpar(col = "white", lwd = 0.5),
                        heatmap_legend_param = list(
                          title = "scaled log2(TPM+1)", 
                          #title_position = "leftcenter-rot",
                          direction = "horizontal")
)

ht2 <- ComplexHeatmap::Heatmap(matrix_notch_expression,
                        cluster_rows = FALSE,
                        cluster_columns = FALSE,
                       # column_split  = col_split,
                        row_title = "Expression of\nNotch\nReceptors",
                        row_title_rot = 0,
                        row_title_gp = gpar(fontsize = 10),
                      # top_annotation = col_ha,
                        show_column_names = TRUE,
                        col = col_fun,
                        rect_gp = gpar(col = "white", lwd = 0.5),
                        show_heatmap_legend = FALSE,
                        heatmap_legend_param = list(
                          title = "scaled log2(TPM+1)", 
                          title_position = "leftcenter-rot")
)

mutation_names <- unique(unlist(matrix_notch_mutation))
mutation_names <- mutation_names[!is.na(mutation_names)]

col_mut = c("lightgrey",sample(merck_colors_ext(1:40))[1:(length(mutation_names)-1)])
names(col_mut) <- mutation_names
col_mut <- col_mut[order(names(col_mut))]

ht3 <- ComplexHeatmap::Heatmap(matrix_notch_mutation,
                        cluster_rows = FALSE,
                        cluster_columns = FALSE,
                       # column_split  = col_split,
                        row_title = "Mutations in\nNotch\nReceptors",
                        row_title_rot = 0,
                        row_title_gp = gpar(fontsize = 10),
                        column_names_gp = gpar(fontsize = 10),
                        show_column_names = TRUE,
                        col = col_mut,
                        na_col = "white",
                        rect_gp = gpar(col = "white", lwd = 0.5),
                        heatmap_legend_param = list(
                          title = "NOTCH Pathway\nMutations")
                        
)

draw(ht %v% ht2 %v% ht3, heatmap_legend_side = "right", annotation_legend_side = "right", merge_legends = TRUE)

```

# Calculation of the signature scores in the experiments

```{r calculation_of_signature_scores, message=FALSE, warning=FALSE, include=FALSE}

signature_score <- function(x, df){
  df %>% 
    dplyr::filter(id_gene_symbol %in% x) %>% 
    dplyr::group_by(sample_name, stripped_cell_line_name, condition_treatment, condition_time, condition_subtype, condition_replicate, condition_batch) %>% 
    dplyr::summarise(quant_signature = mean(quant_scaled_tpm, na.rm = TRUE)) %>% 
    dplyr::ungroup()
}

sig.scores_published.notch <- map_dfr(gs.list.published_notch, ~signature_score(.x, df = tpm_df_scaled), .id = "id_geneset_name")
sig.scores_14.notch <- map_dfr(gene_set_list_14_notch, ~signature_score(.x, df = tpm_df_scaled), .id = "id_geneset_name")
sig.scores_opt.notch <- map_dfr(gs.list.opt_notch, ~signature_score(.x, df = tpm_df_scaled), .id = "id_geneset_name")

all_signatures <- sig.scores_published.notch %>% 
    dplyr::bind_rows(sig.scores_opt.notch)

##### calculate separately the NOTCH Core signature

NOTCH_CORE_SIG_genes <- list(NOTCH_CORE_SIG = c("NOTCH1", "NOTCH2", "NOTCH3", "NOTCH4", "DLL1", "DLL3", "DLL4", "JAG1", "JAG2")) 

sig.notch_core_all_lines <- map_dfr(NOTCH_CORE_SIG_genes, ~signature_score(.x, df = tpm_df_scaled), .id = "id_geneset_name") %>% 
  dplyr::distinct()

NOTCH_CORE_SIG_genes_all_lines <- tpm_df_scaled %>% 
  dplyr::filter(id_gene_symbol %in% !!unlist(NOTCH_CORE_SIG_genes))


## write signatures into excel list:

all_signatures_export <- all_signatures %>% 
    dplyr::bind_rows(sig.notch_core_all_lines) %>% 
  dplyr::mutate(condition_batch = ifelse(condition_batch == "Training", "Basal-like 6 cell line panel", "Breast Cancer 19 cell line panel"))

wb <- createWorkbook()
addWorksheet(wb, "Cell Lines")
writeData(wb, 
          "Cell Lines", 
          all_signatures_export)
saveWorkbook(wb, "../results/signature_scores.xlsx", overwrite = TRUE)

```

```{r cell_lines_ground_state, fig.height=4, fig.width=8}

signature_opt_notch_core_correlation <- sig.notch_core_all_lines %>% 
  bind_rows(sig.scores_opt.notch) %>% 
  dplyr::mutate(id_geneset_name = str_replace_all(id_geneset_name, "\\_", " ")) %>% 
  pivot_wider(names_from = "id_geneset_name",
              values_from = "quant_signature")

signature_opt_notch_core_correlation %>% 
  dplyr::filter(condition_treatment == "FC") %>% 
  ggplot(aes(y = `NOTCH CORE SIG`,
             x = `Notch 20-Gene Signature`)) +
  geom_point(aes(color = stripped_cell_line_name)) +
  scale_color_manual(values = merck_colors_ext(1:20)) +
  geom_smooth(method = "lm",
              color = merck_colors[1]) +
  labs(title = "Cell Lines at Ground State") + 
  stat_cor(method = "pearson", label.x = -0.5, label.y = 2,
           p.accuracy = 0.001, r.accuracy = 0.01,
           size = 4) +
  facet_wrap(~condition_batch)

```

```{r cell_lines_all_states, fig.width=8, fig.height=4}

signature_opt_notch_core_correlation %>% 
  dplyr::mutate(condition_treatment = ifelse(condition_treatment == "FC + DAPT", "DAPT", condition_treatment)) %>% 
  ggplot(aes(y = `NOTCH CORE SIG`,
             x = `Notch 20-Gene Signature`)) +
  geom_point(aes(color = stripped_cell_line_name)) +
  scale_color_manual(values = merck_colors_ext(1:20)) +
  geom_smooth(method = "lm",
              color = merck_colors[1]) +
  labs(title = "Cell Lines at Ground State") + 
  stat_cor(method = "pearson", label.x = -0.5, label.y = 2,
           p.accuracy = 0.001, r.accuracy = 0.01,
           size = 3) +
    facet_grid(condition_batch ~ condition_treatment)
```


```{r correlation_core_all_signatures, fig.height=42, fig.width=9}
# get R values for all, heatmap

cell_lines_core_signature_correlation <- sig.notch_core_all_lines %>% 
  dplyr::rename(`NOTCH CORE SIG` = quant_signature) %>% 
  dplyr::select(-id_geneset_name) %>% 
  dplyr::left_join(sig.scores_opt.notch %>% 
    dplyr::bind_rows(sig.scores_published.notch)) %>% 
  dplyr::mutate(id_geneset_name = str_replace_all(id_geneset_name, "\\_", " ")) %>% 
  dplyr::filter(condition_treatment == "FC")

cell_lines_core_signature_correlation_padj <- cell_lines_core_signature_correlation %>%
  group_by(id_geneset_name, condition_batch) %>%
 cor_test(`NOTCH CORE SIG`, quant_signature) %>% 
  rstatix::adjust_pvalue() %>% 
  rstatix::add_significance() %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(p_level = ifelse(p.adj < 0.001, "<0.001", 
                                 ifelse(p.adj > 0.05, "= ns", 
                                        paste0("= ", signif(p.adj, digits = 2)))))


cell_lines_core_signature_correlation %>% 
  ggplot(aes(y = `NOTCH CORE SIG`,
             x = quant_signature)) +
  facet_grid(id_geneset_name ~ condition_batch) +
  geom_point(aes(color = stripped_cell_line_name),
             size = 1) +
  scale_color_manual(values = merck_colors_ext(1:20)) +
  geom_smooth(method = "lm",
              color = merck_colors[1]) +
  labs(title = "NOTCH Signature corelation to NOTCH Core Signature",
       subtitle = "In Ground State Cell Lines",
       x = "Signature Score",
       color = "Cell Line Name") + 
  geom_text(data = cell_lines_core_signature_correlation_padj,
            aes(y = 1,
                x = -1,
                label = paste("R =", signif(cor, digits = 2), "; p", p_level)),
            size = 2.5,
            vjust = 0.5,
            hjust = 0) +
  theme(strip.text.y = element_text(size = 6, angle = 0, hjust = 0),
        strip.background.y =element_rect(fill = NA,
                                         color = NA))
```

```{r correlation_core_focus_signatures, fig.height=15, fig.width=7}


focus_signatures <- c("Notch 20-Gene Signature", sort(c( 
"Braune_2016_TNBC_Notch_sig_CSL_KO",
"Castel_2013_C2C12_Dll1_vs_DAPT_targets_24h_up",
"WP_NOTCH_SIGNALING_PATHWAY",
"GOBP_NOTCH_SIGNALING_INVOLVED_IN_HEART_DEVELOPMENT",
"Labban_2018_SCC_RBPJ_targets",
"Leontovich_2018_NOTCH3_nuclear_reprogramming_BRCA",
"PID_NOTCH_PATHWAY",
"Stoeck2014_TNBC_ACC_GSI_response_1",
"WP_CANONICAL_AND_NONCANONICAL_NOTCH_SIGNALING",
"KEGG_NOTCH_SIGNALING_PATHWAY",
"Poulsen_2018_N1_repressed",
"HALLMARK_NOTCH_SIGNALING",
"BIOCARTA_NOTCH_PATHWAY",
"VILIMAS_NOTCH1_TARGETS_UP",
"GOBP_POSITIVE_REGULATION_OF_NOTCH_SIGNALING_PATHWAY",
"Xu_2017_MDAMB231_RBPJ_targets")))

labels_keep <- str_replace_all(focus_signatures, "\\_", " ")

cell_lines_core_signature_correlation %>% 
  dplyr::filter(id_geneset_name %in% labels_keep) %>% 
  ggplot(aes(y = `NOTCH CORE SIG`,
             x = quant_signature)) +
  facet_grid(id_geneset_name ~ condition_batch) +
  geom_point(aes(color = stripped_cell_line_name),
             size = 1) +
  scale_color_manual(values = merck_colors_ext(1:20)) +
  geom_smooth(method = "lm",
              color = merck_colors[1]) +
  labs(title = "NOTCH Signature corelation to NOTCH Core Signature",
       subtitle = "In Ground State Cell Lines",
       x = "Signature Score",
       color = "Cell Line Name") + 
  geom_text(data = (cell_lines_core_signature_correlation_padj %>% 
              dplyr::filter(id_geneset_name %in% labels_keep)),
            aes(y = 1,
                x = -1,
                label = paste("R =", signif(cor, digits = 2), "; p", p_level)),
            size = 2.5,
            vjust = 0.5,
            hjust = 0) +
  theme(strip.text.y = element_text(size = 6, angle = 0, hjust = 0),
        strip.background.y =element_rect(fill = NA,
                                         color = NA))
```

```{r correlation_core_signature_heatmap_all, fig.height=8, fig.width=12}

core_signature_correlation_heatmap <- cell_lines_core_signature_correlation_padj %>% 
  dplyr::select(id_geneset_name, cor, condition_batch) %>% 
  pivot_wider(names_from = "condition_batch",
              values_from = "cor") %>% 
  dplyr::mutate(id_geneset_name = str_replace_all(id_geneset_name, "\n", " ")) %>% 
  column_to_rownames("id_geneset_name") %>% t()

core_signature_correlation_heatmap_padj <- cell_lines_core_signature_correlation_padj %>% 
  dplyr::select(id_geneset_name, p.adj, condition_batch) %>% 
  pivot_wider(names_from = "condition_batch",
              values_from = "p.adj") %>% 
  dplyr::mutate(id_geneset_name = str_replace_all(id_geneset_name, "\n", " ")) %>% 
  column_to_rownames("id_geneset_name") %>% t()

core_signature_correlation_heatmap_class <- core_signature_correlation_heatmap_padj

core_signature_correlation_heatmap_class[core_signature_correlation_heatmap_padj > 0.05] <- "ns"
#core_signature_correlation_heatmap_class[core_signature_correlation_heatmap_padj < 0.01] <- "**"
#core_signature_correlation_heatmap_class[core_signature_correlation_heatmap_padj < 0.001] <- "***"

col_fun = colorRamp2(c(-1, 0, 1), c(merck_colors[1], "white", merck_colors[2]))

ComplexHeatmap::Heatmap(core_signature_correlation_heatmap, 
                        name = "mat", 
                        rect_gp = gpar(col = "white", lwd = 2),
                        column_title = "Correlation of NOTCH signatures in breast cancer cell lines\nat ground state to the NOTCH Core Signature",
                        column_names_gp = gpar(fontsize = 8,
                                               fontface = ifelse(grepl("Notch 20", colnames(core_signature_correlation_heatmap)), "bold.italic", "plain")),
                        column_names_max_height = max_text_width(
                          colnames(core_signature_correlation_heatmap), 
                          gp = gpar(fontsize = 8)
                          ),
                        height = unit(5, "mm")*nrow(core_signature_correlation_heatmap),
                        heatmap_legend_param = list(
                          title = "Correlation Coefficient", at = c(-1, 0, 1)
                          ),
                        col = col_fun,
                        cell_fun = function(j, i, x, y, width, height, fill) {
                          if(core_signature_correlation_heatmap_padj[i, j] > 0.05){
                            grid.text(core_signature_correlation_heatmap_class[i, j], x, y, gp = gpar(fontsize = 6))
                          }
                        })


```

```{r correlation_core_signature_heatmap_focus, fig.width=8, fig.height=7}
# signatures to show:


ComplexHeatmap::Heatmap(core_signature_correlation_heatmap[,colnames(core_signature_correlation_heatmap) %in% labels_keep], 
                        name = "mat", 
                        rect_gp = gpar(col = "white", lwd = 2),
                        column_title = "Correlation of NOTCH signatures in breast cancer cell lines\nat ground state to the NOTCH Core Signature",
                        column_names_gp = gpar(fontsize = 8,
                                               fontface = ifelse(grepl("Notch 20", colnames(core_signature_correlation_heatmap)), "bold.italic", "plain")),
                        column_names_max_height = max_text_width(
                          colnames(core_signature_correlation_heatmap), 
                          gp = gpar(fontsize = 8)
                          ),
                        height = unit(5, "mm")*nrow(core_signature_correlation_heatmap),
                        heatmap_legend_param = list(
                          title = "Correlation Coefficient", at = c(-1, 0, 1)
                          ),
                        col = col_fun,
                        cell_fun = function(j, i, x, y, width, height, fill) {
                          if(core_signature_correlation_heatmap_padj[,colnames(core_signature_correlation_heatmap_padj) %in% labels_keep][i, j] > 0.05){
                            grid.text(core_signature_correlation_heatmap_class[,colnames(core_signature_correlation_heatmap_class) %in% labels_keep][i, j], x, y, gp = gpar(fontsize = 7))
                          }
                        })


## save data as excel

xlsx::write.xlsx(cell_lines_core_signature_correlation_padj,
                 file = "../results/Table_S8_cor_signatures_to_core_signature.xlsx",
                 sheet = "cell_line_panel")

```


## ROC-analysis

We calculate the relative signature score (= Condition - Ground State) calculated on the 19 cell line set with known NOTCH alteration levels (as treated with either DAPT or cultivated on immobilized Jagged1). Subsequently, we do a ROC analysis to evaluate accuracy of the signatures to evaluate NOTCH activation levels.  

### top 3 signatures in all cancer subtypes for all timepoints (8h + 72h)

```{r fig.width=9, fig.height=3}

# calculate signature scores for each of the models at each of the timepoints and treatment types

data_signature <- sig.scores_14.notch %>% 
  dplyr::mutate(id_geneset_name = str_replace_all(id_geneset_name, "\\_", " ")) %>% 
  dplyr::filter(condition_batch == "Validation") %>% 
  tidyr::pivot_wider(names_from = "id_geneset_name",
                     values_from = "quant_signature") %>% 
  dplyr::mutate(condition_treatment=dplyr::recode_factor(
    condition_treatment,
    `FC`   = 'ground state',
    `FC + DAPT`      = 'Notch off',
    `Jagged1`     = 'Notch on',
    `Jagged1 + DAPT`  = 'Notch on/off')) %>%
  dplyr::filter(condition_treatment != 'Notch on/off') %>% 
  dplyr::mutate(condition_treatment = droplevels(condition_treatment)) %>%
  tidyr::pivot_longer(-c(1:7),
                      names_to = "id_geneset_name",
                      values_to = "score") %>%
  dplyr::select(-sample_name) %>% 
  tidyr::pivot_wider(names_from = "condition_treatment",
                     values_from = "score") %>%
  dplyr::mutate(`Notch on rel` = `Notch on` - `ground state`,
                  `Notch off rel` = `Notch off` - `ground state`) %>%
  dplyr::select(-`ground state`) %>% 
  tidyr::pivot_longer(-c(1:6),
                      names_to = "condition_treatment",
                      values_to = "quant_rel") %>% 
  dplyr::filter(!is.na(quant_rel))

# Relative scores in all cancer types:

data_signature_rel <- data_signature %>% 
  dplyr::filter(grepl("rel$",condition_treatment),
                id_geneset_name %in% c("signature 5", "signature 9", "signature 11"))

plot_list_3_rel <- list()
conf_int_3_rel <- list()

# response: TRUE labels
# predictor: Signature score

for(i in unique(data_signature_rel$id_geneset_name)){
  
  data_signature_rel_i <- data_signature_rel %>% 
    dplyr::filter(id_geneset_name == !!i)
  
  r <- roc(data_signature_rel_i$condition_treatment,
           data_signature_rel_i$quant_rel,
           levels=c("Notch on rel", "Notch off rel"),
           direction = "auto")
  
  plot_list_3_rel[[i]] <- ggroc(r) +
    labs(title = i,
         y = "Sensitivity (TPR)",
         x = "Specificity (FPR)") +
    geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "grey", linetype = "dashed") +
    scale_y_continuous() +
    scale_x_reverse() +
    annotate("text", 
             x = 0.95,
             hjust = 0,
             y = 0.05, 
             label = paste0("AUC = ", round(auc(r), 2), " (", round(ci.auc(r)[1],2), " - ", round(ci.auc(r)[3],2), ")"))

  conf_int_3_rel[[i]] <- data.frame(id_geneset_name = i,
                                  AUC = as.numeric(auc(r)),
                                  upper = as.numeric(ci.auc(r)[3]),
                                  lower = as.numeric(ci.auc(r)[1]))
}

conf_int_3_rel <- bind_rows(conf_int_3_rel)
DT::datatable(conf_int_3_rel,
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')
            )
          ) 


plot_grid(plotlist=plot_list_3_rel,
          nrow= 1)



```


### top 3 signatures in all cancer subtypes at 8h

```{r fig.width=9, fig.height=3}

# calculate signature scores for each of the models at each of the timepoints and treatment types

data_signature_rel <- data_signature %>% 
  dplyr::filter(grepl("rel$",condition_treatment),
                condition_time == "8h",
                id_geneset_name %in% c("signature 5", "signature 9", "signature 11"))

plot_list_3_8h_rel <- list()
conf_int_3_8h_rel <- list()

# response: TRUE labels
# predictor: Signature score

for(i in unique(data_signature_rel$id_geneset_name)){
  
  data_signature_rel_i <- data_signature_rel %>% 
    dplyr::filter(id_geneset_name == !!i)
  
  r <- roc(data_signature_rel_i$condition_treatment,
           data_signature_rel_i$quant_rel,
           levels=c("Notch on rel", "Notch off rel"),
           direction = "auto")
  
  plot_list_3_8h_rel[[i]] <- ggroc(r) +
    labs(title = i,
         y = "Sensitivity (TPR)",
         x = "Specificity (FPR)") +
    geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "grey", linetype = "dashed") +
    scale_y_continuous() +
    scale_x_reverse() +
    annotate("text", 
             x = 0.95,
             hjust = 0,
             y = 0.05, 
             label = paste0("AUC = ", round(auc(r), 2), " (", round(ci.auc(r)[1],2), " - ", round(ci.auc(r)[3],2), ")"))

  conf_int_3_8h_rel[[i]] <- data.frame(id_geneset_name = i,
                                  AUC = as.numeric(auc(r)),
                                  upper = as.numeric(ci.auc(r)[3]),
                                  lower = as.numeric(ci.auc(r)[1]))
}

conf_int_3_8h_rel <- bind_rows(conf_int_3_8h_rel)
DT::datatable(conf_int_3_8h_rel,
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')
            )
          ) 


plot_grid(plotlist=plot_list_3_8h_rel,
          nrow= 1)



```


### top 3 signatures in all cancer subtypes at 72h

```{r fig.width=9, fig.height=3}

# calculate signature scores for each of the models at each of the timepoints and treatment types

data_signature_rel <- data_signature %>% 
  dplyr::filter(grepl("rel$",condition_treatment),
                condition_time == "72h",
                id_geneset_name %in% c("signature 5", "signature 9", "signature 11"))

plot_list_3_72h_rel <- list()
conf_int_3_72h_rel <- list()

# response: TRUE labels
# predictor: Signature score

for(i in unique(data_signature_rel$id_geneset_name)){
  
  data_signature_rel_i <- data_signature_rel %>% 
    dplyr::filter(id_geneset_name == !!i)
  
  r <- roc(data_signature_rel_i$condition_treatment,
           data_signature_rel_i$quant_rel,
           levels=c("Notch on rel", "Notch off rel"),
           direction = "auto")
  
  plot_list_3_72h_rel[[i]] <- ggroc(r) +
    labs(title = i,
         y = "Sensitivity (TPR)",
         x = "Specificity (FPR)") +
    geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "grey", linetype = "dashed") +
    scale_y_continuous() +
    scale_x_reverse() +
    annotate("text", 
             x = 0.95,
             hjust = 0,
             y = 0.05, 
             label = paste0("AUC = ", round(auc(r), 2), " (", round(ci.auc(r)[1],2), " - ", round(ci.auc(r)[3],2), ")"))

  conf_int_3_72h_rel[[i]] <- data.frame(id_geneset_name = i,
                                  AUC = as.numeric(auc(r)),
                                  upper = as.numeric(ci.auc(r)[3]),
                                  lower = as.numeric(ci.auc(r)[1]))
}

conf_int_3_72h_rel <- bind_rows(conf_int_3_72h_rel)
DT::datatable(conf_int_3_72h_rel,
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')
            )
          ) 


plot_grid(plotlist=plot_list_3_72h_rel,
          nrow= 1)



```

#### AUC comparisons

```{r fig.height=3.2, fig.width=8.5}

conf_int_3 <- conf_int_3_rel %>% 
  dplyr::mutate(facet = "Notch On vs. Notch Off: 8h+72h") %>% 
  bind_rows(conf_int_3_8h_rel  %>%
              dplyr::mutate(facet = "Notch On vs. Notch Off: 8h")) %>% 
  bind_rows(conf_int_3_72h_rel  %>%
              dplyr::mutate(facet = "Notch On vs. Notch Off: 72h"))


conf_int_3 %>% 
    dplyr::mutate(id_geneset_name = factor(id_geneset_name, levels = str_sort(unique(.$id_geneset_name), numeric = TRUE))) %>% 
    dplyr::mutate(facet = factor(facet, levels = c("Notch On vs. Notch Off: 8h+72h",
                                                   "Notch On vs. Notch Off: 8h",
                                                   "Notch On vs. Notch Off: 72h"))) %>% 
  ggplot(aes(x = id_geneset_name,
             y = AUC,
             fill = AUC)) + 
  facet_wrap(~facet) +
  geom_bar(stat="identity") +
  geom_errorbar( aes(ymin=lower, 
                     ymax=upper), 
                 width=0.5, colour="black", alpha=0.9, size=0.5) +
  scale_fill_gradient(low = merck_colors[1],
                      high = merck_colors[6],
                      limits = c(0.5,1),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90) +
  labs(y = "AUC",
       x = "ID",
       title = "ROC Analysis for published signatures",
       subtitle = "19 BC cell line data set, AUC with 95% CI")


```



### top 3 signatures in basal-like

```{r fig.width=9, fig.height=3}

# calculate signature scores for each of the models at each of the timepoints and treatment types
# Relative scores in basal like breast cancer

data_signature_rel <- data_signature %>% 
  dplyr::filter(grepl("rel$",condition_treatment),
                condition_subtype == "basal",
                id_geneset_name %in% c("signature 5", "signature 9", "signature 11"))

plot_list_3_basal_rel <- list()
conf_int_3_basal_rel <- list()

# response: TRUE labels
# predictor: Signature score

for(i in unique(data_signature_rel$id_geneset_name)){
  
  data_signature_rel_i <- data_signature_rel %>% 
    dplyr::filter(id_geneset_name == !!i)
  
  r <- roc(data_signature_rel_i$condition_treatment,
           data_signature_rel_i$quant_rel,
           levels=c("Notch on rel", "Notch off rel"),
           direction = "auto")
  
  plot_list_3_basal_rel[[i]] <- ggroc(r) +
    labs(title = i,
         y = "Sensitivity (TPR)",
         x = "Specificity (FPR)") +
    geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "grey", linetype = "dashed") +
    scale_y_continuous() +
    scale_x_reverse() +
    annotate("text", 
             x = 0.95,
             hjust = 0,
             y = 0.05, 
             label = paste0("AUC = ", round(auc(r), 2), " (", round(ci.auc(r)[1],2), " - ", round(ci.auc(r)[3],2), ")"))

  conf_int_3_basal_rel[[i]] <- data.frame(id_geneset_name = i,
                                  AUC = as.numeric(auc(r)),
                                  upper = as.numeric(ci.auc(r)[3]),
                                  lower = as.numeric(ci.auc(r)[1]))
}

conf_int_3_basal_rel <- bind_rows(conf_int_3_basal_rel)
DT::datatable(conf_int_3_basal_rel,
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')
            )
          ) 

plot_grid(plotlist=plot_list_3_basal_rel,
          nrow= 1)

```



### 14 Signatures in all subtypes (8h+72h)

```{r fig.width=4, fig.height=12}

# Relative scores in all cancer types:

data_signature_rel <- data_signature %>% 
  dplyr::filter(grepl("rel$",condition_treatment))

plot_list_14_rel <- list()
conf_int_14_rel <- list()

# response: TRUE labels
# predictor: Signature score

for(i in unique(data_signature_rel$id_geneset_name)){
  
  data_signature_rel_i <- data_signature_rel %>% 
    dplyr::filter(id_geneset_name == !!i)
  
  r <- roc(data_signature_rel_i$condition_treatment,
           data_signature_rel_i$quant_rel,
           levels=c("Notch on rel", "Notch off rel"),
           direction = "auto")
  
  plot_list_14_rel[[i]] <- ggroc(r) +
    labs(title = i,
         y = "Sensitivity (TPR)",
         x = "Specificity (FPR)") +
    geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "grey", linetype = "dashed") +
    scale_y_continuous() +
    scale_x_reverse() +
    annotate("text", 
             size = 2.5,
             x = 0.95,
             hjust = 0,
             y = 0.05, 
             label = paste0("AUC = ", round(auc(r), 2), " (", round(ci.auc(r)[1],2), " - ", round(ci.auc(r)[3],2), ")"))

  conf_int_14_rel[[i]] <- data.frame(id_geneset_name = i,
                                  AUC = as.numeric(auc(r)),
                                  upper = as.numeric(ci.auc(r)[3]),
                                  lower = as.numeric(ci.auc(r)[1]))
}

conf_int_14_rel <- bind_rows(conf_int_14_rel)
DT::datatable(conf_int_14_rel,
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')
            )
          )  


plot_grid(plotlist=plot_list_14_rel,
          nrow= 7)


xlsx::write.xlsx(conf_int_14_rel,
                 append = TRUE,
                 file = "Signatures_AUCs.xlsx",
                 sheet = "ROC Validation Signatures 8+72h",
                 row.names = FALSE)

                 
```


### 14 Signatures in all subtypes (8h)

```{r fig.width=4, fig.height=12}

# Relative scores in all cancer types:

data_signature_rel <- data_signature %>% 
  dplyr::filter(condition_time == "8h",
                grepl("rel$",condition_treatment))

plot_list_14_8h_rel <- list()
conf_int_14_8h_rel <- list()

# response: TRUE labels
# predictor: Signature score

for(i in unique(data_signature_rel$id_geneset_name)){
  
  data_signature_rel_i <- data_signature_rel %>% 
    dplyr::filter(id_geneset_name == !!i)
  
  r <- roc(data_signature_rel_i$condition_treatment,
           data_signature_rel_i$quant_rel,
           levels=c("Notch on rel", "Notch off rel"),
           direction = "auto")
  
  plot_list_14_8h_rel[[i]] <- ggroc(r) +
    labs(title = i,
         y = "Sensitivity (TPR)",
         x = "Specificity (FPR)") +
    geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "grey", linetype = "dashed") +
    scale_y_continuous() +
    scale_x_reverse() +
    annotate("text", 
             size = 2.5,
             x = 0.95,
             hjust = 0,
             y = 0.05, 
             label = paste0("AUC = ", round(auc(r), 2), " (", round(ci.auc(r)[1],2), " - ", round(ci.auc(r)[3],2), ")"))

  conf_int_14_8h_rel[[i]] <- data.frame(id_geneset_name = i,
                                  AUC = as.numeric(auc(r)),
                                  upper = as.numeric(ci.auc(r)[3]),
                                  lower = as.numeric(ci.auc(r)[1]))
}

conf_int_14_8h_rel <- bind_rows(conf_int_14_8h_rel)
DT::datatable(conf_int_14_8h_rel,
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')
            )
          )  


plot_grid(plotlist=plot_list_14_8h_rel,
          nrow= 7)


xlsx::write.xlsx(conf_int_14_8h_rel,
                 append = TRUE,
                 file = "Signatures_AUCs.xlsx",
                 sheet = "ROC Validation Signatures 8h",
                 row.names = FALSE)
                 
```


### 14 Signatures in all subtypes (72h)

```{r fig.width=4, fig.height=12}

# Relative scores in all cancer types:

data_signature_rel <- data_signature %>% 
  dplyr::filter(condition_time == "72h",
                grepl("rel$",condition_treatment))

plot_list_14_72h_rel <- list()
conf_int_14_72h_rel <- list()

# response: TRUE labels
# predictor: Signature score

for(i in unique(data_signature_rel$id_geneset_name)){
  
  data_signature_rel_i <- data_signature_rel %>% 
    dplyr::filter(id_geneset_name == !!i)
  
  r <- roc(data_signature_rel_i$condition_treatment,
           data_signature_rel_i$quant_rel,
           levels=c("Notch on rel", "Notch off rel"),
           direction = "auto")
  
  plot_list_14_72h_rel[[i]] <- ggroc(r) +
    labs(title = i,
         y = "Sensitivity (TPR)",
         x = "Specificity (FPR)") +
    geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "grey", linetype = "dashed") +
    scale_y_continuous() +
    scale_x_reverse() +
    annotate("text", 
             size = 2.5,
             x = 0.95,
             hjust = 0,
             y = 0.05, 
             label = paste0("AUC = ", round(auc(r), 2), " (", round(ci.auc(r)[1],2), " - ", round(ci.auc(r)[3],2), ")"))

  conf_int_14_72h_rel[[i]] <- data.frame(id_geneset_name = i,
                                  AUC = as.numeric(auc(r)),
                                  upper = as.numeric(ci.auc(r)[3]),
                                  lower = as.numeric(ci.auc(r)[1]))
}

conf_int_14_72h_rel <- bind_rows(conf_int_14_72h_rel)
DT::datatable(conf_int_14_72h_rel,
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')
            )
          )  


plot_grid(plotlist=plot_list_14_72h_rel,
          nrow= 7)


xlsx::write.xlsx(conf_int_14_72h_rel,
                 append = TRUE,
                 file = "Signatures_AUCs.xlsx",
                 sheet = "ROC Validation Signatures 72h",
                 row.names = FALSE)
                 
```

### 14 Signatures in basal-like BRCA

```{r fig.width=4, fig.height=12}

# Relative scores in all cancer types:

data_signature_rel <- data_signature %>% 
  dplyr::filter(condition_subtype == "basal",
                grepl("rel$",condition_treatment))

plot_list_14_basal_rel <- list()
conf_int_14_basal_rel <- list()

# response: TRUE labels
# predictor: Signature score

for(i in unique(data_signature_rel$id_geneset_name)){
  
  data_signature_rel_i <- data_signature_rel %>% 
    dplyr::filter(id_geneset_name == !!i)
  
  r <- roc(data_signature_rel_i$condition_treatment,
           data_signature_rel_i$quant_rel,
           levels=c("Notch on rel", "Notch off rel"),
           direction = "auto")
  
  plot_list_14_basal_rel[[i]] <- ggroc(r) +
    labs(title = i,
         y = "Sensitivity (TPR)",
         x = "Specificity (FPR)") +
    geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "grey", linetype = "dashed") +
    scale_y_continuous() +
    scale_x_reverse() +
    annotate("text", 
             size = 2.5,
             x = 0.95,
             hjust = 0,
             y = 0.05, 
             label = paste0("AUC = ", round(auc(r), 2), " (", round(ci.auc(r)[1],2), " - ", round(ci.auc(r)[3],2), ")"))

  conf_int_14_basal_rel[[i]] <- data.frame(id_geneset_name = i,
                                  AUC = as.numeric(auc(r)),
                                  upper = as.numeric(ci.auc(r)[3]),
                                  lower = as.numeric(ci.auc(r)[1]))
}

conf_int_14_basal_rel <- bind_rows(conf_int_14_basal_rel)
DT::datatable(conf_int_14_basal_rel,
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')
            )
          )  


plot_grid(plotlist=plot_list_14_basal_rel,
          nrow= 7)
```

### Our signature in comparison to other signatures in all subtypes

#### 8h+72h

```{r fig.width=12, fig.height=18}

data_signatures_all <- all_signatures %>% 
  dplyr::mutate(id_geneset_name = str_replace_all(id_geneset_name, "\\_", " ")) %>% 
  dplyr::filter(condition_batch == "Validation") %>% 
  tidyr::pivot_wider(names_from = "id_geneset_name",
                     values_from = "quant_signature") %>% 
  dplyr::mutate(condition_treatment=dplyr::recode_factor(
    condition_treatment,
    `FC`   = 'ground state',
    `FC + DAPT`      = 'Notch off',
    `Jagged1`     = 'Notch on',
    `Jagged1 + DAPT`  = 'Notch on/off')) %>%
  dplyr::filter(condition_treatment != 'Notch on/off') %>% 
  dplyr::mutate(condition_treatment = droplevels(condition_treatment)) %>%
  tidyr::pivot_longer(-c(1:7),
                      names_to = "id_geneset_name",
                      values_to = "score") %>%
  dplyr::select(-sample_name) %>% 
  tidyr::pivot_wider(names_from = "condition_treatment",
                     values_from = "score") %>%
  dplyr::mutate(`Notch on rel` = `Notch on` - `ground state`,
                  `Notch off rel` = `Notch off` - `ground state`) %>%
  dplyr::select(-`ground state`) %>% 
  tidyr::pivot_longer(-c(1:6),
                      names_to = "condition_treatment",
                      values_to = "quant_rel") %>% 
  dplyr::filter(!is.na(quant_rel))

# Relative scores in all cancer types:

data_signature_rel <- data_signatures_all %>% 
  dplyr::filter(grepl("rel$",condition_treatment))

plot_list_all_rel <- list()
conf_int_all_rel <- list()

# response: TRUE labels
# predictor: Signature score

for(i in unique(data_signature_rel$id_geneset_name)){
  
  data_signature_rel_i <- data_signature_rel %>% 
    dplyr::filter(id_geneset_name == !!i)
  
  r <- roc(data_signature_rel_i$condition_treatment,
           data_signature_rel_i$quant_rel,
           levels=c("Notch on rel", "Notch off rel"),
           direction = "auto")
  
  plot_list_all_rel[[i]] <- ggroc(r) +
    labs(title = i,
         y = "Sensitivity (TPR)",
         x = "Specificity (FPR)") +
    geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "grey", linetype = "dashed") +
    scale_y_continuous() +
    scale_x_reverse() +
    theme(title = element_text(size = 6)) +
    annotate("text", 
             size = 2.5,
             x = 0.95,
             hjust = 0,
             y = 0.05, 
             label = paste0("AUC = ", round(auc(r), 2), " (", round(ci.auc(r)[1],2), " - ", round(ci.auc(r)[3],2), ")"))

  conf_int_all_rel[[i]] <- data.frame(id_geneset_name = i,
                                  AUC = as.numeric(auc(r)),
                                  upper = as.numeric(ci.auc(r)[3]),
                                  lower = as.numeric(ci.auc(r)[1]))
}

conf_int_all_rel <- bind_rows(conf_int_all_rel)

DT::datatable(conf_int_all_rel,
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')
            )
          )  

plot_grid(plotlist=plot_list_all_rel,
          nrow= 10)

xlsx::write.xlsx(conf_int_all_rel,
                 file = "Signatures_AUCs.xlsx",
                 append = TRUE,
                 sheet = "ROC Validation Published 8+72h",
                 row.names = FALSE)

```


#### 8h

```{r fig.width=12, fig.height=18}

data_signature_rel <- data_signatures_all %>% 
  dplyr::filter(condition_time == "8h",
                grepl("rel$",condition_treatment))

plot_list_all_8h_rel <- list()
conf_int_all_8h_rel <- list()

# response: TRUE labels
# predictor: Signature score

for(i in unique(data_signature_rel$id_geneset_name)){
  
  data_signature_rel_i <- data_signature_rel %>% 
    dplyr::filter(id_geneset_name == !!i)
  
  r <- roc(data_signature_rel_i$condition_treatment,
           data_signature_rel_i$quant_rel,
           levels=c("Notch on rel", "Notch off rel"),
           direction = "auto")
  
  plot_list_all_8h_rel[[i]] <- ggroc(r) +
    labs(title = i,
         y = "Sensitivity (TPR)",
         x = "Specificity (FPR)") +
    geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "grey", linetype = "dashed") +
    scale_y_continuous() +
    scale_x_reverse() +
    theme(title = element_text(size = 6)) +
    annotate("text", 
             size = 2.5,
             x = 0.95,
             hjust = 0,
             y = 0.05, 
             label = paste0("AUC = ", round(auc(r), 2), " (", round(ci.auc(r)[1],2), " - ", round(ci.auc(r)[3],2), ")"))

  conf_int_all_8h_rel[[i]] <- data.frame(id_geneset_name = i,
                                  AUC = as.numeric(auc(r)),
                                  upper = as.numeric(ci.auc(r)[3]),
                                  lower = as.numeric(ci.auc(r)[1]))
}

conf_int_all_8h_rel <- bind_rows(conf_int_all_8h_rel)

DT::datatable(conf_int_all_8h_rel,
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')
            )
          )  

plot_grid(plotlist=plot_list_all_8h_rel,
          nrow= 10)

xlsx::write.xlsx(conf_int_all_8h_rel,
                 file = "Signatures_AUCs.xlsx",
                 append = TRUE,
                 sheet = "ROC Validation Published 8h",
                 row.names = FALSE)

```

#### 72h

```{r fig.width=12, fig.height=18}

data_signature_rel <- data_signatures_all %>% 
  dplyr::filter(condition_time == "72h",
                grepl("rel$",condition_treatment))

plot_list_all_72h_rel <- list()
conf_int_all_72h_rel <- list()

# response: TRUE labels
# predictor: Signature score

for(i in unique(data_signature_rel$id_geneset_name)){
  
  data_signature_rel_i <- data_signature_rel %>% 
    dplyr::filter(id_geneset_name == !!i)
  
  r <- roc(data_signature_rel_i$condition_treatment,
           data_signature_rel_i$quant_rel,
           levels=c("Notch on rel", "Notch off rel"),
           direction = "auto")
  
  plot_list_all_72h_rel[[i]] <- ggroc(r) +
    labs(title = i,
         y = "Sensitivity (TPR)",
         x = "Specificity (FPR)") +
    geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "grey", linetype = "dashed") +
    scale_y_continuous() +
    scale_x_reverse() +
    theme(title = element_text(size = 6)) +
    annotate("text", 
             size = 2.5,
             x = 0.95,
             hjust = 0,
             y = 0.05, 
             label = paste0("AUC = ", round(auc(r), 2), " (", round(ci.auc(r)[1],2), " - ", round(ci.auc(r)[3],2), ")"))

  conf_int_all_72h_rel[[i]] <- data.frame(id_geneset_name = i,
                                  AUC = as.numeric(auc(r)),
                                  upper = as.numeric(ci.auc(r)[3]),
                                  lower = as.numeric(ci.auc(r)[1]))
}

conf_int_all_72h_rel <- bind_rows(conf_int_all_72h_rel)

DT::datatable(conf_int_all_72h_rel,
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')
            )
          )  

plot_grid(plotlist=plot_list_all_72h_rel,
          nrow= 10)

xlsx::write.xlsx(conf_int_all_72h_rel,
                 file = "Signatures_AUCs.xlsx",
                 append = TRUE,
                 sheet = "ROC Validation Published 72h",
                 row.names = FALSE)

```


#### Visualization as barplot

```{r fig.height=12, fig.width=15}
conf_int_all <- conf_int_all_rel %>% 
  dplyr::mutate(facet = "Notch On vs. Notch Off: 8h+72h") %>% 
  bind_rows(conf_int_all_8h_rel  %>%
              dplyr::mutate(facet = "Notch On vs. Notch Off: 8h")) %>% 
  bind_rows(conf_int_all_72h_rel  %>%
              dplyr::mutate(facet = "Notch On vs. Notch Off: 72h"))


conf_int_all_order <- conf_int_all %>% 
  dplyr::filter(facet == "Notch On vs. Notch Off: 8h+72h") %>% 
  arrange(AUC) %>% 
  pull(id_geneset_name)

conf_int_all$id_geneset_name <- factor(conf_int_all$id_geneset_name, 
                                           levels = conf_int_all_order)

conf_int_all %>% 
    dplyr::mutate(facet = factor(facet, levels = c("Notch On vs. Notch Off: 8h+72h",
                                                   "Notch On vs. Notch Off: 8h",
                                                   "Notch On vs. Notch Off: 72h"))) %>% 
  ggplot(aes(y = id_geneset_name,
             x = AUC,
             fill = AUC)) + 
  facet_wrap(~facet) +
  geom_bar(stat="identity") +
  geom_errorbar( aes(xmin=lower, 
                     xmax=upper), 
                 width=0.5, colour="black", alpha=0.9, size=0.5) +
  scale_fill_gradient(low = merck_colors[1],
                      high = merck_colors[6],
                      limits = c(0.5,1),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90) +
  labs(x = "AUC",
       y = "ID",
       title = "ROC Analysis for published signatures",
       subtitle = "19 BC cell line data set, AUC with 95% CI")
```


```{r fig.height=12, fig.width=15}


conf_int_all_order_desc <- conf_int_all %>% 
  dplyr::filter(facet == "Notch On vs. Notch Off: 8h+72h") %>% 
  arrange(desc(AUC)) %>% 
  pull(id_geneset_name)

conf_int_all$id_geneset_name <- factor(conf_int_all$id_geneset_name, 
                                           levels = conf_int_all_order_desc)

conf_int_all %>% 
    dplyr::mutate(facet = factor(facet, levels = c("Notch On vs. Notch Off: 8h+72h",
                                                   "Notch On vs. Notch Off: 8h",
                                                   "Notch On vs. Notch Off: 72h"))) %>% 
  ggplot(aes(x = id_geneset_name,
             y = AUC,
             fill = AUC)) + 
  facet_wrap(~facet, ncol = 1) +
  geom_bar(stat="identity") +
  geom_errorbar( aes(ymin=lower, 
                     ymax=upper), 
                 width=0.5, colour="black", alpha=0.9, size=0.5) +
  scale_fill_gradient(low = merck_colors[1],
                      high = merck_colors[6],
                      limits = c(0.5,1),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90) +
  labs(y = "AUC",
       x = "ID",
       title = "ROC Analysis for published signatures",
       subtitle = "19 BC cell line data set, AUC with 95% CI")

```

#### Focus signatures

##### 8h + 72h

```{r fig.width=8, fig.height=9}
focus_signatures_space <- str_replace_all(focus_signatures, "\\_", " ")

plot_grid(plotlist=plot_list_all_rel[focus_signatures_space],
          nrow= 5)
```

##### 8h

```{r fig.width=8, fig.height=9}

plot_grid(plotlist=plot_list_all_8h_rel[focus_signatures_space],
          nrow= 5)
```

##### 72h

```{r fig.width=8, fig.height=9}

plot_grid(plotlist=plot_list_all_72h_rel[focus_signatures_space],
          nrow= 5)
```

#### Focus Signatures Visualization as barplot

```{r fig.height=6, fig.width=13}

conf_int_all_rel_focus <- conf_int_all_rel %>% 
  dplyr::mutate(facet = "Notch On vs. Notch Off: 8h+72h") %>% 
  bind_rows(conf_int_all_8h_rel  %>%
              dplyr::mutate(facet = "Notch On vs. Notch Off: 8h")) %>% 
  bind_rows(conf_int_all_72h_rel  %>%
              dplyr::mutate(facet = "Notch On vs. Notch Off: 72h")) %>% 
  dplyr::filter(id_geneset_name %in% focus_signatures_space)


conf_int_all_rel_focus_order <- conf_int_all_rel_focus %>% 
  dplyr::filter(facet == "Notch On vs. Notch Off: 8h+72h") %>% 
  arrange(AUC) %>% 
  pull(id_geneset_name)

conf_int_all_rel_focus$id_geneset_name <- factor(conf_int_all_rel_focus$id_geneset_name, 
                                           levels = conf_int_all_rel_focus_order)


conf_int_all_rel_focus %>% 
    dplyr::mutate(facet = factor(facet, levels = c("Notch On vs. Notch Off: 8h+72h",
                                                   "Notch On vs. Notch Off: 8h",
                                                   "Notch On vs. Notch Off: 72h"))) %>% 
  ggplot(aes(y = id_geneset_name,
             x = AUC,
             fill = AUC)) + 
  facet_wrap(~facet) +
  geom_bar(stat="identity") +
  geom_errorbar( aes(xmin=lower, 
                     xmax=upper), 
                 width=0.5, colour="black", alpha=0.9, size=0.5) +
  scale_fill_gradient(low = merck_colors[1],
                      high = merck_colors[6],
                      limits = c(0.5,1),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90) +
  labs(x = "AUC",
       y = "ID",
       title = "ROC Analysis for published signatures",
       subtitle = "19 BC cell line data set, AUC with 95% CI")

```

### Our signature in comparison to other signatures in basal-like BRCA

```{r fig.width=12, fig.height=18}

# Relative scores in all cancer types:

data_signature_rel <- data_signatures_all %>% 
  dplyr::filter(condition_subtype == "basal",
                grepl("rel$",condition_treatment))

plot_list_all_basal_rel <- list()
conf_int_all_basal_rel <- list()

# response: TRUE labels
# predictor: Signature score

for(i in unique(data_signature_rel$id_geneset_name)){
  
  data_signature_rel_i <- data_signature_rel %>% 
    dplyr::filter(id_geneset_name == !!i)
  
  r <- roc(data_signature_rel_i$condition_treatment,
           data_signature_rel_i$quant_rel,
           levels=c("Notch on rel", "Notch off rel"),
           direction = "auto")
  
  plot_list_all_basal_rel[[i]] <- ggroc(r) +
    labs(title = i,
         y = "Sensitivity (TPR)",
         x = "Specificity (FPR)") +
    geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "grey", linetype = "dashed") +
    scale_y_continuous() +
    scale_x_reverse() +
    theme(title = element_text(size = 6)) +
    annotate("text", 
             size = 2.5,
             x = 0.95,
             hjust = 0,
             y = 0.05, 
             label = paste0("AUC = ", round(auc(r), 2), " (", round(ci.auc(r)[1],2), " - ", round(ci.auc(r)[3],2), ")"))

  conf_int_all_basal_rel[[i]] <- data.frame(id_geneset_name = i,
                                  AUC = as.numeric(auc(r)),
                                  upper = as.numeric(ci.auc(r)[3]),
                                  lower = as.numeric(ci.auc(r)[1]))
}

conf_int_all_basal_rel <- bind_rows(conf_int_all_basal_rel)

DT::datatable(conf_int_all_basal_rel,
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel')
            )
          )  

plot_grid(plotlist=plot_list_all_basal_rel,
          nrow= 10)
```


#### Visualization as barplot

```{r fig.height=10, fig.width=10}

conf_int_all_basal_rel$id_geneset_name <- factor(conf_int_all_basal_rel$id_geneset_name, 
                                           levels = conf_int_all_basal_rel$id_geneset_name[order(conf_int_all_basal_rel$AUC,decreasing = FALSE)])

ggplot(conf_int_all_basal_rel) +
  geom_bar( aes(y=id_geneset_name, 
                x=AUC,
                fill = AUC), 
            stat="identity") +
  geom_errorbar( aes(y=id_geneset_name, xmin=lower, xmax=upper), 
                 width=0.8, colour="black", alpha=0.9, size=0.5) +
  scale_fill_gradient(low = merck_colors[1],
                      high = merck_colors[6],
                      limits = c(0.5,1),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90) +
  labs(y = "Signatures",
       x = "AUC",
       title = "ROC Analysis for published signatures",
       subtitle = "Basal like cell lines of the 19 BC cell line data set, AUC with 95% CI")

```

#### Focus signatures

```{r fig.width=8, fig.height=9}

plot_grid(plotlist=plot_list_all_basal_rel[focus_signatures_space],
          nrow= 5)

```


#### Focus Signatures Visualization as barplot

```{r fig.height=4, fig.width=8}

conf_int_all_basal_rel_focus <- conf_int_all_basal_rel %>% 
  dplyr::filter(id_geneset_name %in% focus_signatures_space) %>% 
  dplyr::mutate(id_geneset_name = droplevels(id_geneset_name))

ggplot(conf_int_all_basal_rel_focus) +
  geom_bar( aes(y=id_geneset_name, 
                x=AUC,
                fill = AUC), 
            stat="identity") +
  geom_errorbar( aes(y=id_geneset_name, xmin=lower, xmax=upper), 
                 width=0.8, colour="black", alpha=0.9, size=0.5) +
  scale_fill_gradient(low = merck_colors[1],
                      high = merck_colors[6],
                      limits = c(0.5,1),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90) +
  labs(y = "Signatures",
       x = "AUC",
       title = "ROC Analysis for published signatures",
       subtitle = "Basal like cell lines of the 19 BC cell line data set, AUC with 95% CI")

```

## Ground State signature scores in the validation cohort of breast cancer cell lines

### Statistics

* basal-like subtype against the rest for the signature comparison
* all comparisons for the plots (only basal against the rest printed) 
  
```{r}

data_signature <- all_signatures %>%
  dplyr::filter(condition_batch == "Validation",
                condition_treatment == "FC") %>%
  dplyr::mutate(condition_treatment=dplyr::recode_factor(
    condition_treatment,
    `FC`   = 'ground state'))

# Pairwise comparisons

pwc_quant.signatures_all <- data_signature %>% 
  #dplyr::group_by(id_geneset_name, condition_time) %>% # time not needed here
  dplyr::group_by(id_geneset_name) %>%
  rstatix::wilcox_test(quant_signature  ~ condition_subtype) %>% 
  adjust_pvalue(method = "BH")  %>% 
  add_significance() %>% 
  dplyr::rename(var1 = ".y.") %>% 
  dplyr::mutate(var2 = "subtype")

data_signature$condition_subtype[data_signature$condition_subtype == "her2"] <- "other"
data_signature$condition_subtype[data_signature$condition_subtype == "luma"] <- "other"
data_signature$condition_subtype[data_signature$condition_subtype == "lumb"] <- "other"

data_signature$condition_subtype <- factor(data_signature$condition_subtype)
data_signature$condition_time <- factor(data_signature$condition_time, levels = c("8h", "72h"))

# Pairwise comparisons

pwc_quant.signatures <- data_signature %>% 
  #dplyr::group_by(id_geneset_name, condition_time) %>% # time not needed here
  dplyr::group_by(id_geneset_name) %>%
  rstatix::wilcox_test(quant_signature  ~ condition_subtype) %>% 
  adjust_pvalue(method = "BH")  %>% 
  add_significance() %>% 
  dplyr::rename(var1 = ".y.") %>% 
  dplyr::mutate(var2 = "subtype")

DT::datatable(pwc_quant.signatures %>% 
                dplyr::select(-statistic),
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel'),        # definition of download buttons; Possible values: 'copy', 'csv', 'excel', 'pdf', 'print', each one increases the file size
            headerCallback = DT::JS(                                    # font-size of table contents
              "function(thead) {",
              "  $(thead).css('font-size', '0.7em');",
              "}"
              )
            )
          )  %>%
  DT::formatStyle(columns = c(1:(ncol(pwc_quant.signatures)-1)), fontSize = '70%')   # reduces font size of column titles

```

```{r, fig.width=4, fig.height=5}

detach("package:coin", unload=TRUE)

for(i in focus_signatures){

  data_signature <- all_signatures %>%
    dplyr::mutate(id_geneset_name = make.names(id_geneset_name)) %>% 
    tidyr::pivot_wider(names_from = "id_geneset_name",
                       values_from = "quant_signature") %>% 
    dplyr::filter(condition_batch == "Validation",
                  condition_treatment == "FC") %>%
    dplyr::mutate(condition_treatment=dplyr::recode_factor(
      condition_treatment,
      `FC`   = 'ground state')) %>% 
    dplyr::filter(condition_subtype != "lumb") # only one cell line of this subtype
    
  data_signature$condition_subtype <- factor(data_signature$condition_subtype, levels = c("basal", "her2", "luma"))
 
  # Pairwise comparisons "dummy calculation" -> use p adjusted from previous calculation
  
  pwc <- data_signature %>% 
    wilcox_test(as.formula(paste(make.names(i), "condition_subtype", sep = " ~ ")), 
              p.adjust.method = "BH") %>% 
   # dplyr::filter(group1 == "basal") %>% 
    add_xy_position(x = "condition_subtype")
  
  pwc_quant.signatures_use <- pwc_quant.signatures_all %>% 
    dplyr::filter(id_geneset_name == !!i)
  
  # ,group1 == "basal"
  
  pwc$p <- pwc_quant.signatures_use$p
  pwc$p.adj <- signif(pwc_quant.signatures_use$p.adj, digits = 3)
  pwc$p.adj.signif <- pwc_quant.signatures_use$p.adj.signif
  
  p <- ggboxplot(data_signature, 
            x = "condition_subtype", 
            y = make.names(i),
            fill = "condition_subtype") + 
    stat_pvalue_manual(
      pwc, 
      label = "p.adj.signif", 
      hide.ns = TRUE) +
  labs(title = paste0("Ground state of signature\n", i),
       x = "Treatment",
       y = "Ground State Signature Score",
       fill = "Subtype",
       caption = get_pwc_label(pwc_quant.signatures)) +
    # ylim(c(-1,1)) +
    scale_fill_manual(values = all_subtype_colors[levels(factor(data_signature$condition_subtype))]) + 
    theme(legend.position="right")
    
  print(p)
}
```

## Relative signature scores (Treatment - Ground State) at both time points in the validation cohort of breast cancer cell lines

```{r, fig.width=4, fig.height=5}

# calculate relative signature score

data_signature_rel <- all_signatures %>%
  dplyr::mutate(id_geneset_name = make.names(id_geneset_name)) %>% 
  tidyr::pivot_wider(names_from = "id_geneset_name",
                     values_from = "quant_signature") %>% 
  dplyr::filter(condition_batch == "Validation") %>%
  dplyr::mutate(condition_treatment=dplyr::recode_factor(
    condition_treatment,
    `FC`   = 'ground state',
    `FC + DAPT`      = 'Notch off',
    `Jagged1`     = 'Notch on',
    `Jagged1 + DAPT`  = 'Notch on/off')) %>% 
  dplyr::filter(condition_subtype != "lumb")
  
complete_cell_lines <- data_signature_rel %>%
  dplyr::filter(condition_treatment != 'Notch on/off') %>% 
  dplyr::mutate(condition_treatment = droplevels(condition_treatment)) %>% 
    dplyr::group_by(stripped_cell_line_name, condition_time) %>%
    dplyr::summarise(n = n()) %>%
    dplyr::ungroup() %>%
    dplyr::filter(n == 3) %>%
    dplyr::select(stripped_cell_line_name, condition_time)

data_signature_rel <- data_signature_rel %>%
  dplyr::right_join(complete_cell_lines) %>%
  tidyr::pivot_longer(-c(1:7),
                      names_to = "id_geneset_name",
                      values_to = "score") %>%
  dplyr::select(-sample_name) %>% 
  tidyr::pivot_wider(names_from = "condition_treatment",
                     values_from = "score") %>%
    dplyr::mutate(`Notch on` = `Notch on` - `ground state`,
                  `Notch off` = `Notch off` - `ground state`) %>%
    dplyr::select(-`ground state`, -`Notch on/off`) %>% 
  dplyr::mutate(condition_subtype = factor(condition_subtype))

data_signature_rel <- data_signature_rel %>% 
  tidyr::pivot_longer(-c(1:6),
                      names_to = "condition_treatment",
                      values_to = "quant_rel")
  
# Kruskal-Wallis Test

res.kruskal_rel <- data_signature_rel %>% 
  dplyr::group_by(id_geneset_name, condition_subtype) %>% 
  kruskal_test(as.formula(paste("quant_rel", "condition_treatment", sep = " ~ ")))
  
# Pairwise comparisons
  
pwc_rel <- data_signature_rel %>% 
  group_by(id_geneset_name, condition_subtype) %>%
  wilcox_test(as.formula(paste("quant_rel", "condition_treatment", sep = " ~ "))) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj") %>% 
  add_xy_position(x = "condition_subtype")

# comparison: basal against other

data_signature_rel_other <- data_signature_rel %>% 
  dplyr::mutate(condition_subtype = factor(ifelse(condition_subtype == "basal", "basal", "other"), levels = c("basal", "other")))

pwc_rel_other <- data_signature_rel_other %>% 
  group_by(id_geneset_name, condition_subtype) %>%
  wilcox_test(as.formula(paste("quant_rel", "condition_treatment", sep = " ~ "))) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj") %>% 
  add_xy_position(x = "condition_subtype")
  
for(i in focus_signatures){
  
  data_signature_use <- data_signature_rel %>% 
    dplyr::filter(id_geneset_name == !!make.names(i)) %>% 
    dplyr::mutate(id_geneset_name = make.names(id_geneset_name))
  
  max_y <- pwc_rel %>% 
    dplyr::filter(id_geneset_name %in% !!focus_signatures) %>% pull(y.position) %>% max
  
  p <- ggboxplot(data_signature_use, 
            x = "condition_subtype", 
            y = "quant_rel",
            fill = "condition_treatment") +
   # facet_wrap(~condition_time) +
  stat_pvalue_manual(pwc_rel %>% 
                       dplyr::filter(id_geneset_name == !!make.names(i)), 
                     hide.ns = TRUE) +
  labs(title = paste0("Relative change of signature:\n",i),
       x = "Treatment",
       y = "Relative Signature Score",
       fill = "Treatment",
       #subtitle = get_test_label(res.kruskal, detailed = TRUE),
       caption = get_pwc_label(pwc_rel)) +
    # ylim(c(-1,1)) +
    coord_cartesian(ylim = c(-max_y,max_y)) +
    scale_fill_manual(values = treatment_colors[levels(factor(data_signature_use$condition_treatment))]) + 
    theme(legend.position="right")
    
  print(p)
}
#dev.off()
```

### Statistics

Only basal-like subtype of interest for multiple testing correction.  

```{r}

DT::datatable(pwc_rel %>% 
                dplyr::select(-statistic, -`.y.`, -y.position, -groups, -x, -xmin, -xmax),
          class = 'cell-border stripe', 
          extensions = 'Buttons',                                       # Add buttons for download
          filter = "top",                                               # Add column search for factor variables
          rownames = FALSE,
          options = list(
            scrollX='800px',
            dom = 'Bfrtip',
            buttons = c('excel'),        # definition of download buttons; Possible values: 'copy', 'csv', 'excel', 'pdf', 'print', each one increases the file size
            headerCallback = DT::JS(                                    # font-size of table contents
              "function(thead) {",
              "  $(thead).css('font-size', '0.7em');",
              "}"
              )
            )
          )  %>%
  DT::formatStyle(columns = c(1:10), fontSize = '70%')   # reduces font size of column titles

```

## Ground State Signature Scores and the relative change in signature score

```{r scatter_notch, fig.height=4, fig.width=5}

data_signature <- data_signature_rel %>% 
  dplyr::rename(delta_signature = quant_rel) %>% 
  dplyr::left_join(all_signatures %>% 
                     dplyr::filter(condition_treatment == "FC") %>% 
                     dplyr::rename(ground_level = "quant_signature") %>% 
                     dplyr::select(-sample_name, -condition_treatment) %>% 
                     dplyr::mutate(id_geneset_name = make.names(id_geneset_name)))

for(i in focus_signatures){
    
  data_scatter <- data_signature %>%
    dplyr::filter(id_geneset_name == !!make.names(i))
  
  sp <- ggscatter(data_scatter, 
                  x = "ground_level", 
                  y = "delta_signature",
                  color = "condition_treatment",
                  shape = "condition_subtype",
                  palette = merck_colors[c(2,1)],
                  legend = "right") +
    geom_smooth(aes(color = condition_treatment), 
                method = "lm",
                fill = "lightgrey") +  
    stat_cor(aes(color = condition_treatment),
                  label.x = min(data_scatter$ground_level)*0.9,
                  #label.y = min(data_scatter$delta_signature)*0.9,
                  p.accuracy = 0.001, 
                  r.accuracy = 0.01,
                  show.legend = FALSE) +
      labs(x = "Ground State Signature Score",
           y = "Relative Signature Score Change",
           title = i,
           color = "Treatment",
           shape = "Subtype")
  print(sp)

}
```


## Ground State Signature Scores and the relative change in signature score - only DAPT

```{r scatter_notch_ground_delta, fig.height=4, fig.width=5}

for(i in focus_signatures){
    
  data_scatter <- data_signature %>%
    dplyr::filter(id_geneset_name == !!make.names(i),
                  condition_treatment == "Notch off")

  sp <- ggscatter(data_scatter, 
                  x = "ground_level", 
                  y = "delta_signature",
                  color = "condition_time",
                  shape = "condition_subtype",
                  palette = merck_colors[c(2,1,3,4)],
                  legend = "right") +
    geom_smooth(aes(color = condition_time), 
                method = "lm",
                fill = "lightgrey") +  
    stat_cor(aes(color = condition_time),
                  label.x = min(data_scatter$ground_level)*0.9,
                  #label.y = min(data_scatter$delta_signature)*0.9,
                  p.accuracy = 0.001, 
                  r.accuracy = 0.01,
                  show.legend = FALSE) +
      labs(x = "Ground State Signature Score",
           y = "Relative Signature Score Change\nupon DAPT treatment",
           title = i,
           color = "Time",
           shape = "Subtype")
  print(sp)

}
```

# Signature relationships

We make use of the Jaccard index to understand the similarity of the different signature genes of the various signatures. Both Jaccard as well as the Overlap Coefficient are calculated and the distances plotted as a heatmaps with complete agglomeration.    

Jaccard Index: This is defined as the size of the intersection divided by the size of the union of the two sets.  
J(A,B)=  AB / AB

Overlap Coefficient: This is defined as the size of the intersection divided by the size of the smaller of the two sets.  
O(A,B)= AB / min(A,B)  

**Range:** Both the Jaccard Index and Overlap Coefficient take values between 0 and 1. A value of 0 indicates that the two sets have no elements in common, and a value of 1 indicates that the sets are identical (for Jaccard) or one is a subset of the other (for Overlap Coefficient).

**Sensitivity to Size:**  
Jaccard Index is more sensitive to the size of the union of the sets. If one set is much larger than the other, the Jaccard Index can be low even if the smaller set is entirely contained within the larger set.

Overlap Coefficient is less sensitive to differences in set size. It can yield a high value if a large proportion of the smaller set is contained within the larger set, even if the absolute size of the intersection is small.

**Interpretation:**  
Jaccard Index is often interpreted as a measure of the relative similarity between two sets, accounting for both the shared and unique elements.
Overlap Coefficient is often interpreted as a measure of the degree to which one set is a subset of another.

**Use Cases:**  
Jaccard Index is generally preferred when you want to account for both the similarities and differences between sets, and when the sets are of roughly equal size.  
Overlap Coefficient might be more appropriate when you are specifically interested in quantifying the degree of overlap, and you expect that one set may be a subset of the other or that the set sizes may be imbalanced.  

## Jaccard Index

```{r fig.height=12, fig.width=12}
library(ggdendro)

all_signatures_to_compare <- c(gs.list.opt_notch, gs.list.published_notch)

jaccard_similarity <- function(list1, list2) {
  return(length(intersect(list1, list2)) / length(union(list1, list2)))
}

jaccard_similarity_matrix <- outer(all_signatures_to_compare, all_signatures_to_compare, Vectorize(function(x, y) jaccard_similarity(x, y)))

hclust_jaccard_distance <- hclust(as.dist(1-jaccard_similarity_matrix), method = 'complete')

ggdendro::ggdendrogram(hclust_jaccard_distance,
             rotate = TRUE,
             size = 0.5)

col_fun = colorRamp2(c(0,0.5, 1), col_heatmap)

rownames(jaccard_similarity_matrix) <- str_replace_all(rownames(jaccard_similarity_matrix), "\\_", " ")
colnames(jaccard_similarity_matrix) <- str_replace_all(colnames(jaccard_similarity_matrix), "\\_", " ")

ComplexHeatmap::Heatmap(jaccard_similarity_matrix,
                        row_names_gp = gpar(fontsize = 8),
                        column_names_gp = gpar(fontsize = 8),
                        col = col_fun,
    row_names_max_width = max_text_width(
        rownames(jaccard_similarity_matrix), 
        gp = gpar(fontsize = 12)
    ),
    column_names_max_height = max_text_width(
        colnames(jaccard_similarity_matrix), 
        gp = gpar(fontsize = 12)
    ),
    clustering_method_rows = "complete",
    clustering_method_columns = "complete")



xlsx::write.xlsx(jaccard_similarity_matrix,
                 file = "../results/Table_S5_similarity_scores.xlsx",
                 sheet = "Jaccard Index Matrix",
                 append = FALSE)

```

### Mean Jaccard Similarity Score

```{r fig.height=7, fig.width=12}
### calculate mean similarity:

# Exclude diagonal from mean calculation
mean_similarities <- apply(jaccard_similarity_matrix, 1, function(row) {
  mean(row[-which(row == max(row))])
})

data.frame(ID = names(mean_similarities),
           `Mean similarities` = mean_similarities,
           check.names = FALSE) %>% 
  dplyr::mutate(ID = str_replace_all(ID, "\\_", " "),
                ID = factor(ID, levels = str_replace_all(names(sort(mean_similarities)), "\\_", " ")),
                textlabel = round(`Mean similarities`, digits = 2)) %>% 
  ggplot(aes(x = ID,
             y = `Mean similarities`,
             color = `Mean similarities`,
             label = textlabel)) +
  geom_segment(aes(x = ID,
                   xend = ID,
                   y = 0,
                   yend = `Mean similarities`),
               color = "darkgrey",
               size = 1) +
  geom_point(size = 6,
             shape = 16) +
  geom_text(color = "white",
            fontface = "bold",
            size = 2) +
  scale_color_gradient(low = merck_colors[1],
                      high = merck_colors[6],
                      limits = c(0,max(mean_similarities)),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90) +
  coord_cartesian(ylim = c(0, max(mean_similarities)*1.1)) +
  labs(title = "Mean Jaccard Similarities")


xlsx::write.xlsx(mean_similarities,
                 file = "../results/Table_S5_similarity_scores.xlsx",
                 sheet = "Mean Jaccard Similarity Index",
                 append = TRUE)

```


## Overlap Coefficient

```{r fig.height=12, fig.width=12}

overlap_coefficient <- function(set1, set2) {
  intersection <- length(intersect(set1, set2))
  smaller_set_size <- min(length(set1), length(set2))
  return(intersection / smaller_set_size)
}

# overlap_coefficient(all_signatures_to_compare$`Notch 20-Gene Signature`, all_signatures_to_compare$`Notch 20-Gene Signature`)

overlap_coefficient_matrix <- outer(all_signatures_to_compare, all_signatures_to_compare, Vectorize(function(x, y) overlap_coefficient(x, y)))

hclust_overlap_distance <- hclust(as.dist(1-overlap_coefficient_matrix), method = 'complete')

ggdendro::ggdendrogram(hclust_overlap_distance,
             rotate = TRUE,
             size = 0.5)

col_fun = colorRamp2(c(0,0.5, 1), col_heatmap)

rownames(overlap_coefficient_matrix) <- str_replace_all(rownames(overlap_coefficient_matrix), "\\_", " ")
colnames(overlap_coefficient_matrix) <- str_replace_all(colnames(overlap_coefficient_matrix), "\\_", " ")

ComplexHeatmap::Heatmap(overlap_coefficient_matrix,
                        row_names_gp = gpar(fontsize = 8),
                        column_names_gp = gpar(fontsize = 8),
                        col = col_fun,
    row_names_max_width = max_text_width(
        rownames(overlap_coefficient_matrix), 
        gp = gpar(fontsize = 12)
    ),
    column_names_max_height = max_text_width(
        colnames(overlap_coefficient_matrix), 
        gp = gpar(fontsize = 12)
    ),
    clustering_method_rows = "complete",
    clustering_method_columns = "complete")

xlsx::write.xlsx(overlap_coefficient_matrix,
                 file = "../results/Table_S5_similarity_scores.xlsx",
                 sheet = "Overlap Coefficient Matrix",
                 append = TRUE)

```

## Mean Overlap Coefficient Score

```{r fig.height=7, fig.width=12}
### calculate mean similarity:

# Exclude diagonal from mean calculation
mean_similarities <- apply(overlap_coefficient_matrix, 1, function(row) {
  mean(row[-which(row == max(row))])
})

data.frame(ID = names(mean_similarities),
           `Mean similarities` = mean_similarities,
           check.names = FALSE) %>% 
  dplyr::mutate(ID = str_replace_all(ID, "\\_", " "),
                ID = factor(ID, levels = str_replace_all(names(sort(mean_similarities)), "\\_", " ")),
                textlabel = round(`Mean similarities`, digits = 2)) %>% 
  ggplot(aes(x = ID,
             y = `Mean similarities`,
             color = `Mean similarities`,
             label = textlabel)) +
  geom_segment(aes(x = ID,
                   xend = ID,
                   y = 0,
                   yend = `Mean similarities`),
               color = "darkgrey",
               size = 1) +
  geom_point(size = 6,
             shape = 16) +
  geom_text(color = "white",
            fontface = "bold",
            size = 2) +
  scale_color_gradient(low = merck_colors[1],
                      high = merck_colors[6],
                      limits = c(0,max(mean_similarities)),
                      na.value = merck_colors[1]) +
  rotate_x_text(angle = 90) +
  coord_cartesian(ylim = c(0, max(mean_similarities)*1.1)) +
  labs(title = "Mean Overlap Coefficient")


xlsx::write.xlsx(mean_similarities,
                 file = "../results/Table_S5_similarity_scores.xlsx",
                 sheet = "Mean Overlap Coefficient Score",
                 append = TRUE)


```




```{r}

# save.image("../RData/06-Image.RData")

pwc_delta.signatures_all = pwc_rel
pwc_delta.signatures = pwc_rel_other

save(gs.list.published_notch, 
     gs.list.opt_notch,
     gsea_Jagged1_DAPT,
     gsea_Jagged1_DAPT_8h,
     gsea_Jagged1_DAPT_72h,
     pwc_quant.signatures,
     pwc_quant.signatures_all,
     pwc_delta.signatures,
     pwc_delta.signatures_all,
     file = "../RData/Part6.RData")

```

# References {.unlisted .unnumbered}

<div id="refs"></div>

# Appendix {.unlisted .unnumbered}
  
```{r SessionInfo}
pander::pander(sessionInfo())
```
